---
title: "HAV Batch Analysis Report"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    df_print: paged
    self_contained: true
params:
  batch_dir:    "data/Batch-1"
  batch_fa:     "data/Batch-1/Batch-1.fa"
  dataset_date: "2026-04-10"
  dataset_dir:  "data/local_datasets/2026-04-10"
  out_base:     "data/Batch-1/output"
  n_neighbors:  30
  mode:         "sanger"
  zoom_count: 5
  zoom_width: 35
  min_called_fraction: 0.5
  entropy_window: 15
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 12,
  fig.height = 7
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(ape)
  library(ggtree)
  library(phangorn)
  library(kableExtra)
  library(scales)
  library(patchwork)
  library(jsonlite)
  library(pheatmap)
  library(grid)
})

# ── Paths ────────────────────────────────────────────────────────────────────
proj_dir     <- normalizePath(file.path(dirname(knitr::current_input(dir = TRUE)), ".."),
                              mustWork = FALSE)
#batch_dir    <- file.path(proj_dir, params$batch_dir)
batch_dir    <- file.path(params$batch_dir)
#out_dir      <- file.path(batch_dir, "output")
out_dir      <- file.path(params$out_base)
trees_dir    <- file.path(out_dir, "trees")
#dataset_base <- file.path(proj_dir, "data", "local_datasets", params$dataset_date)
dataset_base <- file.path(params$dataset_dir)
metadata_tsv <- file.path(dataset_base, "metadata_corrected.tsv")
batch_name   <- basename(params$batch_dir)
input_fasta  <- file.path(params$batch_fa)

# ── Helper: safe read_tsv ────────────────────────────────────────────────────
safe_tsv <- function(path, ...) {
  if (!file.exists(path)) return(tibble())
  read_tsv(path, show_col_types = FALSE, ...)
}

safe_csv_sc <- function(path, ...) {
  # NextClade CSV uses semicolons
  if (!file.exists(path)) return(tibble())
  read_delim(path, delim = ";", show_col_types = FALSE, ...)
}

# Safe FASTA reader with validation (prevents ape::read.dna segfault)
safe_read_dna_fasta <- function(path) {
  # Check file existence
  if (!file.exists(path)) return(NULL)
  
  # Check file size (empty file would cause problems)
  if (file.size(path) == 0) return(NULL)
  
  # Quick validation: first line should be a header starting with '>'
  first_line <- readLines(path, n = 1)
  if (length(first_line) == 0 || !grepl("^>", first_line)) {
    warning("Invalid FASTA file (no valid header): ", path)
    return(NULL)
  }
  
  # Count sequences by counting '>' lines
  all_lines <- readLines(path)
  n_seqs <- sum(grepl("^>", all_lines))
  if (n_seqs < 1) return(NULL)
  
  # Try to read with ape::read.dna, but catch errors
  # (tryCatch won't catch segfaults, but at least we tried validation)
  tryCatch({
    ape::read.dna(path, format = "fasta", as.character = TRUE)
  }, error = function(e) {
    warning("Error reading FASTA file: ", path, " - ", e$message)
    NULL
  })
}

# Null-coalescing helper (used in parse_nearest)
`%||%` <- function(a, b) if (!is.null(a) && length(a) > 0) a else b

# Normalize IDs to match FASTA headers: spaces → "_", non-ASCII stripped
norm_id <- function(x) {
  x <- gsub(" ", "_", x)
  iconv(x, from = "UTF-8", to = "ASCII", sub = "")
}

# ── Load data ─────────────────────────────────────────────────────────────────

# Reference metadata
meta <- safe_tsv(metadata_tsv)
if (nrow(meta) > 0 && "date" %in% names(meta)) {
  meta <- meta %>% mutate(date = as.character(date))
}

# Community sequence metadata from hav-vp1-2b-lineages tree.json
# Provides clade_membership + lineage_phylo for NCBI accessions used as neighbors
.ltree_json <- file.path(proj_dir, "data", "nextclade_datasets",
                         "hav-vp1-2b-lineages", "tree.json")
community_meta <- tibble(id = character(),
                         community_clade = character(),
                         community_lineage_phylo = character())
if (file.exists(.ltree_json)) {
  .ltree <- fromJSON(.ltree_json, simplifyDataFrame = FALSE)
  .cl_leaves <- function(node) {
    if (is.null(node$children) || length(node$children) == 0) {
      na <- node$node_attrs %||% list()
      list(tibble(
        id                      = node$name %||% NA_character_,
        community_clade         = na$clade_membership$value %||% NA_character_,
        community_lineage_phylo = na$lineage_phylo$value    %||% NA_character_
      ))
    } else {
      unlist(lapply(node$children, .cl_leaves), recursive = FALSE)
    }
  }
  community_meta <- bind_rows(.cl_leaves(.ltree$tree)) %>%
    mutate(
      community_clade         = str_to_upper(community_clade),
      community_clade         = if_else(is.na(community_clade) | community_clade %in% c("", "UNASSIGNED"),
                                        NA_character_, community_clade),
      community_lineage_phylo = if_else(is.na(community_lineage_phylo) | community_lineage_phylo %in% c("", "unassigned"),
                                        NA_character_, community_lineage_phylo)
    )
  rm(.ltree, .cl_leaves)
}
rm(.ltree_json)

# BLAST results
blast_raw <- safe_tsv(file.path(out_dir, "blast_results.tsv"))

# ── Amplicon coverage + primer scheme (whole-genome / ViroConstrictor only) ───
# ViroConstrictor writes one mean-coverage value per amplicon per sample to
# all_amplicon_coverage.csv. Sanger batches have no such file, so the whole
# amplicon section is skipped downstream when `has_amplicon` is FALSE.
amplicon_cov_path <- file.path(batch_dir, "viroconstrictor", "results",
                               "combined", "all_samples", "all_amplicon_coverage.csv")
has_amplicon  <- file.exists(amplicon_cov_path)
amplicon_long <- tibble()
mincov_val    <- NA_real_

if (has_amplicon) {
  .ac <- readr::read_csv(amplicon_cov_path, show_col_types = FALSE)
  # Columns: Virus, Reference_ID, amplicon_names (= sample name), then one
  # column per amplicon (e.g. HAVIB_001 … HAVIB_006).
  .amp_cols <- setdiff(names(.ac), c("Virus", "Reference_ID", "amplicon_names"))
  if (length(.amp_cols) > 0) {
    amplicon_long <- .ac %>%
      rename(sample = amplicon_names) %>%
      pivot_longer(all_of(.amp_cols), names_to = "amplicon_raw", values_to = "coverage") %>%
      mutate(
        sample   = word(as.character(sample), 1),
        amplicon = as.integer(str_extract(amplicon_raw, "[0-9]+$")),
        coverage = as.numeric(coverage)
      ) %>%
      arrange(sample, amplicon)
  }
}

# Primer scheme table (primer name ↔ amplicon number) from the batch samplesheet
primer_scheme_tbl <- tibble()
.samplesheet_path <- file.path(batch_dir, "samplesheet.tsv")
if (has_amplicon && file.exists(.samplesheet_path)) {
  .ss <- safe_tsv(.samplesheet_path)
  if ("MinCov" %in% names(.ss)) {
    mincov_val <- suppressWarnings(as.numeric(.ss$MinCov[1]))
  }
  if ("Primers" %in% names(.ss)) {
    .primer_rel   <- unique(.ss$Primers)
    .primer_rel   <- .primer_rel[!is.na(.primer_rel) & .primer_rel != "NONE"]
    .primer_paths <- file.path(proj_dir, .primer_rel)
    .primer_paths <- .primer_paths[file.exists(.primer_paths)]
    if (length(.primer_paths) > 0) {
      # Read primer records (header + sequence) preserving file order.
      .read_primer_fa <- function(p) {
        ln  <- readLines(p, warn = FALSE)
        hdr <- which(startsWith(ln, ">"))
        tibble(primer   = sub("^>", "", ln[hdr]),
               sequence = toupper(ln[pmin(hdr + 1L, length(ln))]))
      }
      .renamed_fa <- bind_rows(lapply(.primer_paths, .read_primer_fa))

      # The original (pre-ViroConstrictor) primer names live in a sibling
      # *_original.fasta with identical record order, so pair them positionally.
      .orig_paths <- sub("_renamed\\.fasta$", "_original.fasta", .primer_paths)
      .orig_paths <- .orig_paths[file.exists(.orig_paths)]
      .renamed_fa$original <- NA_character_
      if (length(.orig_paths) > 0) {
        .orig_fa <- bind_rows(lapply(.orig_paths, .read_primer_fa))
        if (nrow(.orig_fa) == nrow(.renamed_fa))
          .renamed_fa$original <- .orig_fa$primer
      }

      # Primer names look like HAVIB_1_LEFT / HAVIB_1_RIGHT — scheme, amplicon, side.
      primer_scheme_tbl <- .renamed_fa %>%
        distinct(primer, .keep_all = TRUE) %>%
        tidyr::extract(primer, c("scheme", "amplicon", "side"),
                       "^(.*)_([0-9]+)_(LEFT|RIGHT)$", remove = FALSE) %>%
        filter(!is.na(amplicon)) %>%
        mutate(amplicon = as.integer(amplicon)) %>%
        select(scheme, amplicon, side, primer, original, sequence) %>%
        pivot_wider(names_from = side,
                    values_from = c(primer, original, sequence)) %>%
        arrange(scheme, amplicon)
    }
  }
}

# ── Genotype helpers ──────────────────────────────────────────────────────────
# HAV genotypes are written in canonical roman form with an uppercase sub-letter
# (IA, IB, IIA, IIIA). Two normalisations are needed:
#   * harmonise_genotype()  — upper-case + trim any genotype label (Ib → IB).
#   * genotype_from_name()  — derive a genotype from a reference name that
#     encodes it in arabic form (e.g. "helgenom_gt_1a" → IA) for hits that carry
#     no genotype in the metadata.
.arabic_to_roman_gt <- c(`1` = "I", `2` = "II", `3` = "III",
                         `4` = "IV", `5` = "V", `6` = "VI", `7` = "VII")

harmonise_genotype <- function(x) {
  x <- str_to_upper(str_trim(as.character(x)))
  x[x %in% c("", "NA")] <- NA_character_
  x
}

genotype_from_name <- function(x) {
  m     <- str_match(str_to_lower(as.character(x)), "gt[_-]?([0-9]+)\\s*([a-z]?)")
  roman <- .arabic_to_roman_gt[m[, 2]]
  ifelse(is.na(roman), NA_character_, paste0(roman, str_to_upper(m[, 3])))
}

# NextClade — lineages dataset (clade + lineage_phylo)
nc_lineages_raw <- safe_tsv(file.path(out_dir, "lineages", "nextclade.tsv"))

# ── Check for NextClade alignment failures ──────────────────────────────────
nc_errors <- tibble()
if (nrow(nc_lineages_raw) > 0 && "errors" %in% names(nc_lineages_raw)) {
  nc_errors <- nc_lineages_raw %>%
    filter(!is.na(errors) & errors != "") %>%
    select(seqName, errors)
  if (nrow(nc_errors) > 0) {
    warning(
      "NextClade alignment failed for ", nrow(nc_errors), " sequence(s).\n",
      "This usually means sequences are too short or don't match the reference dataset.\n",
      "Try running with --min-match-length 10 or lower (default is 10)."
    )
  }
}

nc_lineages <- nc_lineages_raw
if (nrow(nc_lineages) > 0) {
  if (!"lineage_phylo" %in% names(nc_lineages)) {
    nc_lineages <- nc_lineages %>% mutate(lineage_phylo = NA_character_)
  }
  nc_lineages <- nc_lineages %>%
    select(
      seqName,
      clade,
      lineage_phylo,
      totalSubstitutions,
      coverage,
      qc_status = qc.overallStatus,
      qc_score  = qc.overallScore,
      substitutions,
      aaSubstitutions,
      privateNucMuts_total = privateNucMutations.totalPrivateSubstitutions
    ) %>%
    # Ensure coverage is numeric (NextClade may return FALSE for failed alignments)
    mutate(coverage = as.numeric(coverage))
} else {
  nc_lineages <- tibble(
    seqName = character(), clade = character(), lineage_phylo = character(),
    totalSubstitutions = integer(), coverage = double(), qc_status = character(),
    qc_score = double(), substitutions = character(), aaSubstitutions = character(),
    privateNucMuts_total = integer()
  )
}

# Harmonise the NextClade clade label to canonical uppercase (IA, IB, IIIA …).
if (nrow(nc_lineages) > 0) {
  nc_lineages <- nc_lineages %>% mutate(clade = harmonise_genotype(clade))
}

# All query names (from NC-lineages)
all_queries <- nc_lineages$seqName

# Metadata for joining reference info to BLAST hits
meta_for_join <- if (nrow(meta) > 0) {
  meta_base <- meta %>% select(id, any_of(c("date", "genotype", "lineage"))) %>% rename(sseqid = id)
  
  # Conditionally rename columns if they exist
  if ("genotype" %in% names(meta_base)) meta_base <- rename(meta_base, hit_genotype = genotype)
  if ("lineage" %in% names(meta_base)) meta_base <- rename(meta_base, hit_lineage = lineage)
  if ("date" %in% names(meta_base)) meta_base <- rename(meta_base, hit_date = date)
  
  meta_base
} else {
  tibble(sseqid = character())
}

# Only proceed with join if blast_raw has data and the required column exists
blast_top <- if (nrow(blast_raw) > 0 && "sseqid" %in% names(blast_raw)) {
  blast_raw %>%
    left_join(meta_for_join, by = "sseqid") %>%
    arrange(qseqid, desc(bitscore)) %>%
    group_by(qseqid) %>%
    mutate(hit_rank = row_number()) %>%
    ungroup()
} else {
  # Return empty tibble with expected structure if no BLAST results
  blast_raw %>% mutate(hit_rank = NA_integer_)
}

blast_best <- blast_top %>%
  filter(hit_rank == 1) %>%
  select(qseqid, sseqid,
         blast_pident    = pident,
         blast_mismatch  = mismatch,
         blast_qcovs     = qcovs,
         blast_bitscore  = bitscore,
         any_of(c("hit_lineage", "hit_genotype")))

# Ensure hit_lineage and hit_genotype columns exist
if (!"hit_lineage" %in% names(blast_best)) blast_best$hit_lineage <- NA_character_
if (!"hit_genotype" %in% names(blast_best)) blast_best$hit_genotype <- NA_character_

# Harmonise and rename
blast_best <- blast_best %>%
  mutate(
    blast_lineage  = hit_lineage,
    blast_genotype = coalesce(as.character(harmonise_genotype(hit_genotype)),
                              as.character(genotype_from_name(sseqid)))
  ) %>%
  select(-any_of(c("hit_lineage", "hit_genotype")))

# ── Master summary table ──────────────────────────────────────────────────────
# BLAST truncates FASTA headers at the first whitespace, so its qseqid is only
# the first token of the full sequence name (e.g. "barcode84" rather than
# "barcode84 HAV LC128713.1_IB mincov=20"). Join on that first token so WGS
# consensus headers (which carry spaces) still match. For single-word Sanger
# names the first token equals the full name, so this is backward-compatible.
summary_tbl <- nc_lineages %>%
  mutate(.join_key = word(seqName, 1)) %>%
  left_join(blast_best %>% 
            mutate(.join_key = word(qseqid, 1)) %>% 
            select(.join_key, qseqid, everything()),
            by = ".join_key") %>%
  select(-.join_key)

# ── Input FASTA summary (raw batch input) ───────────────────────────────────
read_fasta_simple <- function(path) {
  if (!file.exists(path)) return(tibble(header = character(), sequence = character()))

  lines <- readLines(path, warn = FALSE)
  if (length(lines) == 0) return(tibble(header = character(), sequence = character()))

  recs <- list()
  cur_h <- NULL
  cur_s <- character()

  flush_rec <- function() {
    if (is.null(cur_h)) return(NULL)
    recs[[length(recs) + 1]] <<- tibble(
      header = cur_h,
      sequence = paste(cur_s, collapse = "")
    )
    NULL
  }

  for (ln in lines) {
    ln <- trimws(ln)
    if (!nzchar(ln)) next
    if (startsWith(ln, ">")) {
      flush_rec()
      cur_h <- sub("^>", "", ln)
      cur_s <- character()
    } else {
      cur_s <- c(cur_s, ln)
    }
  }
  flush_rec()

  bind_rows(recs)
}

parse_primer_header <- function(header) {
  parts <- strsplit(header, "[[:space:]]+")[[1]]
  name <- parts[1]
  meta <- list()
  if (length(parts) > 1) {
    for (token in parts[-1]) {
      if (grepl("=", token, fixed = TRUE)) {
        kv <- strsplit(token, "=", fixed = TRUE)[[1]]
        if (length(kv) == 2) meta[[tolower(kv[1])]] <- kv[2]
      }
    }
  }
  list(name = name, meta = meta)
}

iupac_sets <- c(
  A = "A", C = "C", G = "G", T = "T", U = "T",
  R = "AG", Y = "CT", S = "GC", W = "AT", K = "GT", M = "AC",
  B = "CGT", D = "AGT", H = "ACT", V = "ACG", N = "ACGT", I = "ACGT", `5` = "ACGT"
)

primer_rc <- function(seq) {
  seq <- str_to_upper(seq)
  rc_map <- c(A = "T", C = "G", G = "C", T = "A", U = "A",
              R = "Y", Y = "R", S = "S", W = "W", K = "M", M = "K",
              B = "V", D = "H", H = "D", V = "B", N = "N", I = "N", `5` = "N")
  bases <- strsplit(seq, "", fixed = TRUE)[[1]]
  paste0(rev(unname(rc_map[bases])), collapse = "")
}

base_compatible <- function(seq_base, primer_base) {
  seq_base <- toupper(seq_base)
  primer_base <- toupper(primer_base)
  seq_chars <- strsplit(iupac_sets[[seq_base]] %||% seq_base, "", fixed = TRUE)[[1]]
  primer_chars <- strsplit(iupac_sets[[primer_base]] %||% primer_base, "", fixed = TRUE)[[1]]
  length(intersect(seq_chars, primer_chars)) > 0
}

find_best_end_overlap <- function(sequence, primer_sequence, side = c("left", "right"), min_overlap = 8L) {
  side <- match.arg(side)
  seq_chars <- strsplit(str_to_upper(sequence), "", fixed = TRUE)[[1]]
  primer_chars <- strsplit(str_to_upper(primer_sequence), "", fixed = TRUE)[[1]]
  max_overlap <- min(length(seq_chars), length(primer_chars))
  if (max_overlap < min_overlap) return(0L)

  for (overlap in seq(max_overlap, min_overlap, by = -1L)) {
    if (side == "left") {
      seq_window <- seq_chars[seq_len(overlap)]
    } else {
      seq_window <- tail(seq_chars, overlap)
    }
    primer_window <- primer_chars[seq_len(overlap)]
    matched <- vapply(seq_len(overlap), function(i) base_compatible(seq_window[[i]], primer_window[[i]]), logical(1))
    if (all(matched)) return(overlap)
  }
  0L
}

# Primer-like IUPAC symbols commonly found in degenerate primer regions
primer_like_codes <- c("R", "Y", "W", "S", "K", "M", "B", "D", "H", "V", "N")
called_codes <- c("A", "C", "G", "T")

fasta_raw <- read_fasta_simple(input_fasta)

input_seq_summary <- fasta_raw %>%
  mutate(
    seqName = header,
    sequence = str_to_upper(sequence),
    seq_chars = map(sequence, ~strsplit(.x, "", fixed = TRUE)[[1]]),
    length = map_int(seq_chars, length),
    ambiguous_count = map_int(seq_chars, ~sum(!(.x %in% called_codes))),
    ambiguous_chars = map_chr(seq_chars, ~{
      x <- sort(unique(.x[!(.x %in% called_codes)]))
      if (length(x) == 0) "" else paste(x, collapse = ",")
    }),
    primer_like_count = map_int(seq_chars, ~sum(.x %in% primer_like_codes)),
    primer_like_chars = map_chr(seq_chars, ~{
      x <- sort(unique(.x[.x %in% primer_like_codes]))
      if (length(x) == 0) "" else paste(x, collapse = ",")
    })
  ) %>%
  select(seqName, length, ambiguous_count, ambiguous_chars,
         primer_like_count, primer_like_chars, sequence)

# ── Primer catalog and end-overlap scan ──────────────────────────────────────
primer_min_overlap <- 8L
primer_catalog <- read_fasta_simple(file.path(proj_dir, "data", "Primers", "PCR_primers", "primers.fa")) %>%
  mutate(
    parsed = map(header, parse_primer_header),
    primer_name = map_chr(parsed, "name"),
    primer_direction = map_chr(parsed, ~.x$meta$direction %||% ""),
    primer_region = map_chr(parsed, ~.x$meta$region %||% ""),
    primer_comment = map_chr(parsed, ~.x$meta$comment %||% ""),
    primer_sequence = str_to_upper(sequence),
    primer_rc_sequence = map_chr(primer_sequence, primer_rc),
    primer_end = if_else(primer_direction == "F", "5' end", "3' end")
  ) %>%
  select(primer_name, primer_direction, primer_region, primer_comment, primer_sequence, primer_rc_sequence, primer_end)

primer_hits <- if (nrow(input_seq_summary) > 0 && nrow(primer_catalog) > 0) {
  map_dfr(seq_len(nrow(input_seq_summary)), function(i) {
    seq_row <- input_seq_summary[i, ]
    map_dfr(seq_len(nrow(primer_catalog)), function(j) {
      primer_row <- primer_catalog[j, ]
      scan_seq <- if (primer_row$primer_direction == "F") primer_row$primer_sequence[[1]] else primer_row$primer_rc_sequence[[1]]
      left_overlap <- find_best_end_overlap(seq_row$sequence[[1]], scan_seq, side = "left", min_overlap = primer_min_overlap)
      right_overlap <- find_best_end_overlap(seq_row$sequence[[1]], scan_seq, side = "right", min_overlap = primer_min_overlap)
      hits <- tibble(
        seqName = seq_row$seqName[[1]],
        sequence_length = seq_row$length[[1]],
        primer_name = primer_row$primer_name[[1]],
        primer_direction = primer_row$primer_direction[[1]],
        primer_region = primer_row$primer_region[[1]],
        primer_end = primer_row$primer_end[[1]],
        match_side = c("left", "right"),
        overlap_nt = c(left_overlap, right_overlap)
      ) %>%
        filter(overlap_nt >= primer_min_overlap)

      if (nrow(hits) == 0) return(tibble())

      hits %>%
        mutate(
          start = if_else(match_side == "left", 1L, sequence_length - overlap_nt + 1L),
          end = if_else(match_side == "left", overlap_nt, sequence_length),
          matched_sequence = substr(seq_row$sequence[[1]], start, end),
          matched_fraction = round(overlap_nt / nchar(scan_seq), 2),
          end_label = if_else(match_side == "left", "5' end", "3' end")
        )
    })
  })
} else {
  tibble()
}

primer_hits_table <- if (nrow(primer_hits) > 0) {
  primer_hits %>%
    arrange(seqName, match_side, desc(overlap_nt), primer_name)
} else {
  tibble()
}

primer_hit_plot <- if (nrow(primer_hits_table) > 0) {
  plot_df <- primer_hits_table %>%
    mutate(
      seqName = factor(seqName, levels = unique(seqName)),
      primer_label = if_else(primer_region == "", primer_name, paste0(primer_name, " (", primer_region, ")"))
    )

  ggplot(plot_df, aes(y = seqName)) +
    geom_segment(aes(x = start, xend = end, yend = seqName, color = primer_label),
                 linewidth = 5.5, lineend = "round", alpha = 0.9) +
    geom_point(aes(x = start), size = 1.8, color = "#111827") +
    geom_point(aes(x = end), size = 1.8, color = "#111827") +
    facet_wrap(~match_side, ncol = 1, scales = "free_y") +
    scale_x_continuous(expand = expansion(mult = c(0.02, 0.04))) +
    labs(
      title = "Potential primer remnants on input sequences",
      subtitle = paste0("Compatible end-overlaps of at least ", primer_min_overlap, " nt against built-in primer set"),
      x = "Position in input sequence (nt)",
      y = NULL,
      color = "Primer"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom"
    )
} else {
  NULL
}

# ── Primer / input-sequence coverage map ────────────────────────────────────
# Coordinate system: NC_001489.1 full-genome positions.
# The 349 bp NextClade reference maps to 2919-3267 (offset = 2918 to convert
# 1-based 349 bp coords → full-genome coords).
nc_ref_offset    <- 2918L
coverage_map_win <- c(2798L, 3475L)  # display window

# Best primer binding sites (full-genome) from GFF3 analysis
primerbind <- tibble(
  primer_name = c("HAV_6.1", "HAV_8.2", "HAV_11", "HAV_10"),
  fg_start    = c(2832L, 2886L, 3375L, 3417L),
  fg_end      = c(2857L, 2914L, 3403L, 3445L),
  direction   = c("F", "F", "R", "R"),
  role        = c("First-outer", "nested-inner", "nested-inner", "First-outer")
)

# Input sequence positions in full-genome coords
seq_fg <- nc_lineages_raw %>%
  select(seqName, alignmentStart, alignmentEnd) %>%
  mutate(
    fg_start = if_else(!is.na(alignmentStart), as.integer(alignmentStart) + nc_ref_offset, NA_integer_),
    fg_end   = if_else(!is.na(alignmentEnd),   as.integer(alignmentEnd)   + nc_ref_offset, NA_integer_),
    aligned  = !is.na(fg_start)
  )

# Which sequence ends have flagged primer remnants?
left_hit_seqs  <- if (nrow(primer_hits_table) > 0) {
  primer_hits_table %>% filter(match_side == "left")  %>% pull(seqName) %>% unique()
} else {
  character()
}
right_hit_seqs <- if (nrow(primer_hits_table) > 0) {
  primer_hits_table %>% filter(match_side == "right") %>% pull(seqName) %>% unique()
} else {
  character()
}

# Build y positions: reference at bottom, then primers, then input sequences
bar_h     <- 0.37
prim_gap  <- 1.4
ref_y     <- 0
n_prim    <- nrow(primerbind)
prim_ys   <- setNames(seq_len(n_prim) * 1.0 + 0.5, rev(primerbind$primer_name))
seq_names_ord <- rev(seq_fg$seqName)
n_seqs    <- length(seq_names_ord)
seq_ys    <- setNames(seq_along(seq_names_ord) + n_prim + prim_gap, seq_names_ord)

prim_colors <- c(
  "HAV_6.1" = "#1f77b4", "HAV_8.2" = "#2ca02c",
  "HAV_11"  = "#d62728", "HAV_10"  = "#ff7f0e"
)

ref_df   <- tibble(xmin = coverage_map_win[[1]], xmax = coverage_map_win[[2]],
                   ymin = ref_y - bar_h, ymax = ref_y + bar_h)

primerbind_df <- primerbind %>%
  mutate(
    y    = prim_ys[primer_name],
    ymin = y - bar_h, ymax = y + bar_h,
    arw_x    = if_else(direction == "F", fg_end,   fg_start),
    arw_xend = if_else(direction == "F", fg_end + 16L, fg_start - 16L)
  )

seq_df <- seq_fg %>%
  mutate(
    y    = seq_ys[seqName],
    ymin = y - bar_h, ymax = y + bar_h,
    xmin = if_else(aligned, fg_start, coverage_map_win[[1]]),
    xmax = if_else(aligned, fg_end,   coverage_map_win[[2]]),
    has_left_hit  = seqName %in% left_hit_seqs,
    has_right_hit = seqName %in% right_hit_seqs
  )

# Primer-remnant highlight bars at sequence ends (orange overlay)
hit_left_df <- seq_df %>% filter(aligned, has_left_hit) %>%
  mutate(xmin_h = fg_start, xmax_h = fg_start + 29L)
hit_right_df <- seq_df %>% filter(aligned, has_right_hit) %>%
  mutate(xmin_h = fg_end - 29L, xmax_h = fg_end)
hit_overlay_df <- bind_rows(hit_left_df, hit_right_df)

coverage_map_plot <- ggplot() +
  # Product region shading
  annotate("rect", xmin = 2832, xmax = 3445,
           ymin = -Inf, ymax = Inf, fill = "#bbbbbb", alpha = 0.15) +
  annotate("rect", xmin = 2886, xmax = 3403,
           ymin = -Inf, ymax = Inf, fill = "#888888", alpha = 0.15) +
  # Reference bar
  geom_rect(data = ref_df,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "#444444", color = NA, inherit.aes = FALSE) +
  # Aligned input sequence bars
  geom_rect(data = seq_df %>% filter(aligned),
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "#aec6e8", color = "#1a6bb5", linewidth = 0.3, inherit.aes = FALSE) +
  # Unaligned sequences (dashed)
  geom_rect(data = seq_df %>% filter(!aligned),
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "#e8e8e8", color = "#999999", linewidth = 0.3,
            linetype = "dashed", inherit.aes = FALSE) +
  # Primer-remnant overlays at sequence ends
  geom_rect(data = hit_overlay_df,
            aes(xmin = xmin_h, xmax = xmax_h, ymin = ymin, ymax = ymax),
            fill = "#e05c00", alpha = 0.6, color = NA, inherit.aes = FALSE) +
  # Primer binding site bars
  geom_rect(data = primerbind_df,
            aes(xmin = fg_start, xmax = fg_end, ymin = ymin, ymax = ymax,
                fill = primer_name),
            color = "white", linewidth = 0.5, inherit.aes = FALSE) +
  scale_fill_manual(values = prim_colors, name = "Primer") +
  # Direction arrows
  geom_segment(data = primerbind_df,
               aes(x = arw_x, xend = arw_xend, y = y, yend = y,
                   color = primer_name),
               arrow = arrow(length = unit(0.12, "cm"), type = "closed"),
               linewidth = 0.9, inherit.aes = FALSE) +
  scale_color_manual(values = prim_colors, guide = "none") +
  # Axes and labels
  scale_x_continuous(
    limits = coverage_map_win,
    labels = scales::comma,
    expand = expansion(mult = 0),
    name = "NC_001489.1 position (bp)"
  ) +
  scale_y_continuous(
    breaks = c(ref_y, unname(prim_ys), unname(seq_ys)),
    labels = c("Reference", names(prim_ys), names(seq_ys)),
    expand = expansion(add = 0.9)
  ) +
  labs(
    title    = "Primer binding sites and input sequence coverage",
    subtitle = paste0(
      "Shading: outer PCR product (light, ", 2832, "-", 3445, " bp) and ",
      "inner/nested product (dark, ", 2886, "-", 3403, " bp).\n",
      "Orange = detected primer remnant at sequence end. Dashed = failed alignment."
    ),
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y        = element_text(size = 9),
    legend.position    = "right"
  )

# ── Junction-focused alignment view (from Nextclade) ───────────────────────
junction_aln_path <- file.path(out_dir, "lineages", "nextclade.aligned.fasta")
junction_aln <- safe_read_dna_fasta(junction_aln_path)

junction_is_usable <- !is.null(junction_aln) && nrow(junction_aln) >= 2

junction_msa_plot <- NULL
junction_var_plot <- NULL

if (junction_is_usable) {
  junction_ids <- rownames(junction_aln)

  junction_order <- summary_tbl %>%
    select(seqName, clade, lineage_phylo, qc_status) %>%
    mutate(
      clade_rank = dplyr::coalesce(clade, "ZZZ"),
      lineage_rank = dplyr::coalesce(lineage_phylo, "ZZZ"),
      qc_rank = dplyr::coalesce(qc_status, "ZZZ")
    ) %>%
    arrange(clade_rank, lineage_rank, qc_rank, seqName) %>%
    pull(seqName)

  junction_order <- unique(c(junction_order[junction_order %in% junction_ids], junction_ids))

  junction_labels <- summary_tbl %>%
    select(seqName, clade, lineage_phylo) %>%
    mutate(
      clade = if_else(is.na(clade) | clade == "", "?", clade),
      lineage_phylo = if_else(is.na(lineage_phylo) | lineage_phylo == "", "?", lineage_phylo),
      label = paste0(seqName, " [", clade, "/", lineage_phylo, "]")
    ) %>%
    select(seqName, label)

  label_map <- setNames(junction_labels$label, junction_labels$seqName)

  dna_ss_junc <- tryCatch(Biostrings::readDNAStringSet(junction_aln_path), error = function(e) NULL)
  if (!is.null(dna_ss_junc)) {
    keep_ids <- junction_order[junction_order %in% names(dna_ss_junc)]
    if (length(keep_ids) >= 2) {
      dna_ss_junc <- dna_ss_junc[keep_ids]
      names(dna_ss_junc) <- dplyr::coalesce(label_map[names(dna_ss_junc)], names(dna_ss_junc))

      junction_msa_plot <- ggmsa::ggmsa(
        dna_ss_junc,
        start = 1,
        end = Biostrings::width(dna_ss_junc)[1],
        color = "Chemistry_NT",
        seq_name = TRUE,
        font = NULL,
        show.legend = TRUE
      ) +
        labs(
          title = "Nextclade junction alignment (full displayed window)",
          subtitle = "All batch query sequences aligned to the short VP1-2B junction reference",
          x = "Junction alignment position",
          y = NULL
        ) +
        theme_bw(base_size = 9) +
        theme(
          plot.title = element_text(face = "bold", size = 10),
          axis.text.y = element_text(size = 6.5),
          axis.text.x = element_text(size = 7)
        )
    }
  }

  mat_j <- tolower(junction_aln[junction_order, , drop = FALSE])
  called_j <- c("a", "c", "g", "t")

  .cons_j <- function(col) {
    vals <- col[col %in% called_j]
    if (length(vals) == 0) return(NA_character_)
    tab <- sort(table(vals), decreasing = TRUE)
    names(tab)[1]
  }

  consensus_j <- apply(mat_j, 2, .cons_j)
  snp_count_j <- sapply(seq_len(ncol(mat_j)), function(k) {
    col <- mat_j[, k]
    cons <- consensus_j[k]
    if (is.na(cons)) return(0L)
    sum(col %in% called_j & col != cons)
  })
  var_cols_j <- which(snp_count_j > 0)

  if (length(var_cols_j) > 0) {
    mat_var_j <- mat_j[, var_cols_j, drop = FALSE]
    pos_var_j <- var_cols_j
    row_ids_j <- rownames(mat_var_j)

    display_levels_j <- rev(dplyr::coalesce(label_map[row_ids_j], row_ids_j))

    var_grid_j <- expand.grid(
      seq_id = row_ids_j,
      col_i = seq_len(ncol(mat_var_j)),
      KEEP.OUT.ATTRS = FALSE,
      stringsAsFactors = FALSE
    ) %>%
      as_tibble() %>%
      mutate(
        base = mat_var_j[cbind(match(seq_id, row_ids_j), col_i)],
        position = pos_var_j[col_i],
        seq_label = dplyr::coalesce(label_map[seq_id], seq_id),
        base_class = case_when(
          base %in% called_j ~ str_to_upper(base),
          base == "-" ~ "Gap",
          TRUE ~ "Ambig"
        )
      )

    nt_pal_j <- c(
      "A" = "#4daf4a", "C" = "#377eb8", "G" = "#ff7f00", "T" = "#e41a1c",
      "Gap" = "#f5f5f5", "Ambig" = "#6a3d9a"
    )

    junction_var_plot <- ggplot(
      var_grid_j,
      aes(x = position, y = factor(seq_label, levels = display_levels_j), fill = base_class)
    ) +
      geom_tile(color = "white", linewidth = 0.15) +
      scale_fill_manual(values = nt_pal_j, drop = FALSE) +
      labs(
        title = paste0("Nextclade junction variant-only view (", length(var_cols_j), " variable sites)"),
        x = "Junction alignment position (variable sites)",
        y = NULL,
        fill = "Base"
      ) +
      theme_bw(base_size = 10) +
      theme(
        axis.text.y = element_text(size = 7),
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 10),
        legend.position = "bottom"
      )
  }
}

input_overview <- if (nrow(input_seq_summary) > 0) {
  lens <- input_seq_summary$length
  tibble(
    metric = c(
      "Total sequences",
      "Unique headers",
      "Duplicated headers",
      "Min length",
      "Max length",
      "Mean length",
      "Median length",
      "Sequences with ambiguous bases",
      "Total ambiguous bases",
      "Sequences with primer-like IUPAC codes",
      "Total primer-like IUPAC bases"
    ),
    value = c(
      nrow(input_seq_summary),
      n_distinct(input_seq_summary$seqName),
      nrow(input_seq_summary) - n_distinct(input_seq_summary$seqName),
      min(lens),
      max(lens),
      round(mean(lens), 2),
      round(stats::median(lens), 2),
      sum(input_seq_summary$ambiguous_count > 0),
      sum(input_seq_summary$ambiguous_count),
      sum(input_seq_summary$primer_like_count > 0),
      sum(input_seq_summary$primer_like_count)
    )
  )
} else {
  tibble(metric = "Input FASTA", value = "Not found")
}

# Ensure lineage_phylo column exists
if (!"lineage_phylo" %in% names(summary_tbl)) summary_tbl$lineage_phylo <- NA_character_

# Is lineage assigned (not a NODE_ placeholder)?
is_lineage <- function(x) !is.na(x) & !str_starts(x, "NODE_")

summary_tbl <- summary_tbl %>%
  mutate(
    clade_agree_blast = case_when(
      !is.na(blast_genotype) & !is.na(clade) ~
        str_to_upper(replace_na(blast_genotype, "")) ==
          str_to_upper(replace_na(clade, "")),
      TRUE ~ NA
    )
  )

# ── Read samplesheet for LabWareID and SequenceID ─────────────────────────────
.ss_sample_info <- tibble()
.ss_path <- file.path(batch_dir, "auto_samplesheet.tsv")
if (!file.exists(.ss_path)) {
  # Try finding any samplesheet
  .ss_files <- list.files(batch_dir, pattern = "samplesheet\\.tsv$", full.names = TRUE)
  if (length(.ss_files) > 0) .ss_path <- .ss_files[1]
}
if (file.exists(.ss_path)) {
  .ss_sample_info <- safe_tsv(.ss_path) %>%
    select(Sample, any_of(c("LabWareID", "SekvensID"))) %>%
    distinct()
}

# ── Add input sequence length to summary ─────────────────────────────────────
.seq_lengths <- if (nrow(input_seq_summary) > 0) {
  input_seq_summary %>%
    transmute(seqName, input_length = length)
} else {
  tibble(seqName = character(), input_length = integer())
}

# ── Count BLAST database occurrences per lineage and find similar lineages ────
# This helps identify how common the detected lineage is in the reference database
# and warns if other lineages are equally similar (SNP distance <= 3)
.db_lineage_summary <- if (nrow(metadata_tsv_df <- safe_tsv(metadata_tsv)) > 0 && 
                            "community_lineage" %in% names(metadata_tsv_df) &&
                            nrow(blast_raw) > 0) {
  # Count sequences per lineage in database
  lineage_counts <- metadata_tsv_df %>%
    count(community_lineage, name = "db_count") %>%
    rename(blast_lineage = community_lineage)
  
  # Join BLAST results with metadata to get lineage information
  blast_with_lineage <- blast_raw %>%
    left_join(metadata_tsv_df %>% 
              select(id, community_lineage) %>% 
              rename(sseqid = id, lineage = community_lineage),
              by = "sseqid")
  
  # For each query, identify best hit and find alternative lineages within SNP threshold
  similar_lineages <- if (nrow(blast_with_lineage) > 0 && !all(is.na(blast_with_lineage$lineage))) {
    blast_with_lineage %>%
      filter(!is.na(lineage)) %>%  # Only keep rows with lineage info
      group_by(qseqid) %>%
      arrange(mismatch) %>%
      # Get best hit info
      mutate(
        best_rank = row_number(),
        best_sseqid = first(sseqid),
        best_lineage = first(lineage),
        best_mismatch = first(mismatch)
      ) %>%
      # Filter for alternative hits within SNP threshold
      filter(
        mismatch <= best_mismatch + 3,  # Within 3 SNPs of best
        lineage != best_lineage,        # Different lineage
        sseqid != best_sseqid           # Not the best hit itself
      ) %>%
      # Aggregate alternative lineages for this query
      group_by(qseqid, best_sseqid) %>%
      summarise(
        similar_lineages = paste(unique(sort(lineage)), collapse = ", "),
        .groups = "drop"
      ) %>%
      # Keep only one entry per best hit (per query)
      select(qseqid, sseqid = best_sseqid, similar_lineages)
  } else {
    tibble(qseqid = character(), sseqid = character(), similar_lineages = character())
  }
  
  # Combine counts with info about similar lineages
  list(
    counts = lineage_counts,
    similar = similar_lineages
  )
} else {
  list(
    counts = tibble(blast_lineage = character(), db_count = integer()),
    similar = tibble(qseqid = character(), sseqid = character(), similar_lineages = character())
  )
}

.blast_lineage_counts <- .db_lineage_summary$counts %>%
  rename(db_lineage_count = db_count)

.similar_lineages_map <- .db_lineage_summary$similar %>%
  setNames(c("qseqid", "sseqid", "similar_lineages")) %>%
  {if (nrow(.) > 0) . else tibble(sseqid = character(), similar_lineages = character())}

# ── Batch triage: one confidence flag per sample ──────────────────────────────
# Lets the expert skip the obvious calls and focus on the few that need a look.
# The flag is driven by genotype confidence (the report's primary question), not
# by raw BLAST %identity — for WGS the nearest reference rarely reaches
# outbreak-level identity, so a %id threshold would flag everything.
.called <- if (nrow(input_seq_summary) > 0) {
  input_seq_summary %>%
    transmute(join_key   = word(seqName, 1),
              called_frac = if_else(length > 0, (length - ambiguous_count) / length, NA_real_))
} else tibble(join_key = character(), called_frac = numeric())

triage_tbl <- summary_tbl %>%
  mutate(join_key = word(seqName, 1)) %>%
  left_join(.called, by = "join_key") %>%
  left_join(.seq_lengths, by = "seqName") %>%
  left_join(.ss_sample_info, by = c("seqName" = "Sample")) %>%
  left_join(.blast_lineage_counts, by = "blast_lineage") %>%
  left_join(.db_lineage_summary$similar, 
            by = c("qseqid" = "qseqid", "sseqid" = "sseqid")) %>%
  mutate(
    # Format similar lineages warning
    similar_lineages_msg = if_else(!is.na(similar_lineages) & similar_lineages != "",
                                   paste0("⚠ Also similar (≤3 SNPs): ", similar_lineages, " — see heatmap"),
                                   ""),
    flag = case_when(
      is.na(clade)                                   ~ "No call",
      !is.na(clade_agree_blast) & !clade_agree_blast ~ "Review",
      TRUE                                           ~ "Clear"
    ),
    flag_reason = case_when(
      flag == "No call" & !is.na(called_frac) ~ sprintf("No NextClade call (%.0f%% of genome called)", 100 * called_frac),
      flag == "No call"                       ~ "No NextClade call",
      flag == "Review"                        ~ "NextClade and BLAST genotype disagree",
      TRUE                                    ~ ""
    )
  )
```

# `r batch_name` — HAV Analysis Summary

This report compares complementary HAV analyses run on **`r batch_name`**:

| Analysis | Description |
|---|---|
| **NC-lineages** | NextClade with `hav-vp1-2b-lineages` dataset — assigns clade (IA/IB/IIA/IIIA) + sub-genotype `lineage_phylo` |
| **BLAST** | BLAST against local Norwegian HAV sequence database |

---

# Batch overview {.tabset}

## Verdict table

```{r triage-table}
.flag_col <- c(Clear = "#2ca02c", Review = "#d62728", `No call` = "#7f7f7f")
.flag_ord <- c("Review", "No call", "Clear")

triage_tbl %>%
  arrange(factor(flag, levels = .flag_ord), seqName) %>%
  transmute(
    `LabWareID`     = if_else(!is.na(LabWareID), LabWareID, "—"),
    `SequenceID`    = if_else(!is.na(SekvensID), SekvensID, "—"),
    `Genotype (NC)` = if_else(!is.na(clade), clade, "—"),
    `Genotype (BLAST)` = if_else(!is.na(blast_genotype), blast_genotype, "—"),
    `Agree`         = case_when(is.na(clade_agree_blast) ~ "—",
                                clade_agree_blast ~ "✓", TRUE ~ "✗"),
    `BLAST top hit` = if_else(!is.na(sseqid), sseqid, "—"),
    `BLAST lineage` = if_else(!is.na(blast_lineage), blast_lineage, "—"),
    `BLAST SNPs`    = if_else(!is.na(blast_mismatch), as.character(blast_mismatch), "—"),
    `DB lineage count` = if_else(!is.na(db_lineage_count), as.character(db_lineage_count), "—"),
    `Similar BLAST lineage` = if_else(similar_lineages_msg != "", similar_lineages_msg, "—"),
    `Input length`  = if_else(!is.na(input_length), as.character(input_length), "—"),
    `Called %`      = if_else(!is.na(called_frac), percent(called_frac, 1), "—"),
    `Flag`          = cell_spec(flag, color = "white", background = .flag_col[flag], bold = TRUE),
    `Notes`         = flag_reason
  ) -> .triage_display

# Calculate background colors for BLAST SNPs column (same as Per-sequence summary)
.snp_colors <- .triage_display$`BLAST SNPs` %>%
  map_chr(~ {
    if (. == "—") {
      ""  # Empty string for no background
    } else {
      snp_count <- as.numeric(.)
      if (snp_count == 0) {
        "#c8e6c9"  # Light green for 0 SNPs
      } else if (snp_count <= 2) {
        "#fff9c4"  # Light yellow for 1-2 SNPs
      } else {
        ""  # Empty string for no background
      }
    }
  })

.triage_display %>%
  kbl(escape = FALSE,
      caption = "Batch overview — sorted with items needing review first. Green = clear, red = genotype disagreement, grey = no NextClade call (usually low coverage). Called % = percentage of sequence bases that are unambiguous (A/C/G/T; excludes N and other ambiguous codes). DB lineage count = number of sequences in reference database with same BLAST lineage.") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
  column_spec(7, width = "30em", extra_css = "white-space: nowrap; overflow: visible;") %>%  # BLAST lineage - prevent wrapping
  column_spec(8, background = .snp_colors, bold = TRUE)  # Color code BLAST SNPs
```

`r {n_clear <- sum(triage_tbl$flag=="Clear"); n_rev <- sum(triage_tbl$flag=="Review"); n_no <- sum(triage_tbl$flag=="No call"); sprintf("**%d clear · %d to review · %d no call** of %d samples.", n_clear, n_rev, n_no, nrow(triage_tbl))}`

## Placement tree (NextClade)

```{r placement-tree, fig.width=11, fig.height=11}
.nwk_path <- file.path(out_dir, "lineages", "nextclade.nwk")
.ptr <- if (file.exists(.nwk_path)) tryCatch(read.tree(.nwk_path), error = function(e) NULL) else NULL

if (is.null(.ptr)) {
  cat("NextClade placement tree (`nextclade.nwk`) not found — run the NextClade step first.")
} else {
  .qids <- word(all_queries, 1)
  .tipmeta <- tibble(label = .ptr$tip.label) %>%
    left_join(community_meta %>% transmute(label = id, ref_clade = community_clade), by = "label") %>%
    left_join(triage_tbl %>% transmute(label = word(seqName, 1), q_clade = clade), by = "label") %>%
    mutate(
      is_query = label %in% .qids,
      Genotype = coalesce(q_clade, ref_clade),
      Genotype = if_else(is.na(Genotype) | Genotype %in% c("", "NA"), "unassigned", str_to_upper(Genotype))
    )

  .gtypes <- sort(setdiff(unique(.tipmeta$Genotype), "unassigned"))
  .pal <- setNames(scales::hue_pal()(length(.gtypes)), .gtypes)
  .pal["unassigned"] <- "grey80"

  p_place <- ggtree(.ptr, layout = "fan", open.angle = 12, size = 0.15, color = "grey55") %<+% .tipmeta +
    geom_tippoint(aes(color = Genotype, size = is_query, alpha = is_query)) +
    scale_size_manual(values = c(`FALSE` = 0.5, `TRUE` = 3.2), guide = "none") +
    scale_alpha_manual(values = c(`FALSE` = 0.55, `TRUE` = 1), guide = "none") +
    scale_color_manual(values = .pal, name = "Genotype / clade",
                       breaks = .gtypes) +
    ggtree::geom_tiplab2(aes(label = ifelse(is_query, label, NA), color = Genotype),
                         size = 3, fontface = "bold", offset = 0.001, na.rm = TRUE) +
    guides(color = guide_legend(override.aes = list(size = 3))) +
    labs(title = "Batch samples placed on the NextClade reference tree",
         subtitle = sprintf("%d reference tips + %d batch samples (bold red-rim labels). Placement uses the VP1-2B junction.",
                            sum(!.tipmeta$is_query), sum(.tipmeta$is_query))) +
    theme(legend.position = "right",
          plot.title = element_text(face = "bold"))

  print(p_place)
}
```

---

# 0. Input sequence summary {.tabset}

## Coverage map

```{r coverage-map, out.width="100%"}
cov_height <- max(4.5, 1.8 + 0.45 * nrow(seq_fg))
cov_file <- tempfile(fileext = ".png")
ggsave(cov_file, plot = coverage_map_plot, width = 11, height = cov_height, dpi = 150, limitsize = FALSE)
knitr::include_graphics(cov_file)
```

## Junction alignment (focused viewer)

This viewer is intentionally focused on the short Nextclade junction region.
It keeps alignment inspection interpretable even when input and tree alignments
are full-genome.

```{r junction-alignment-viewer, out.width="100%"}
if (!junction_is_usable) {
  cat("Junction alignment file not found or contains fewer than two aligned sequences:", junction_aln_path)
} else {
  if (!is.null(junction_msa_plot)) {
    msa_file <- tempfile(fileext = ".png")
    msa_h <- max(3.2, 0.22 * nrow(junction_aln) + 1.8)
    ggsave(msa_file, plot = junction_msa_plot, width = 16, height = msa_h, dpi = 150, limitsize = FALSE)
    knitr::include_graphics(msa_file)
  }

  if (!is.null(junction_var_plot)) {
    var_file <- tempfile(fileext = ".png")
    var_h <- max(3.2, 0.22 * nrow(junction_aln) + 1.8)
    ggsave(var_file, plot = junction_var_plot, width = 16, height = var_h, dpi = 150, limitsize = FALSE)
    knitr::include_graphics(var_file)
  } else {
    cat("No variable sites were detected in the Nextclade junction alignment.")
  }
}
```

## Overview

```{r input-overview}
input_overview %>%
  kbl(col.names = c("Metric", "Value")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

Primer-like nucleotide codes are reported as IUPAC degenerate symbols:
**R, Y, W, S, K, M, B, D, H, V, N**.

## Potential primer remnants

This scan checks the sequence ends for compatible overlaps against the built-in primer set
and highlights the best matching primer remnant on each end when the overlap is at least
`r primer_min_overlap` nt.

```{r input-primer-hits}
if (nrow(primer_hits_table) > 0) {
  primer_hits_table %>%
    transmute(
      Sequence = seqName,
      `Sequence end` = end_label,
      Primer = primer_name,
      `Primer region` = if_else(primer_region == "", "—", primer_region),
      `Overlap nt` = overlap_nt,
      `Matched fraction` = percent(matched_fraction, accuracy = 1),
      Start = start,
      `End nt` = end,
      `Matched sequence` = matched_sequence
    ) %>%
    kbl(escape = FALSE, caption = "Compatible end-overlaps to built-in primers") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
    scroll_box(width = "100%", height = "320px")

  plot_file <- tempfile(fileext = ".png")
  plot_height <- max(4.5, min(14, 0.55 * n_distinct(primer_hits_table$seqName) + 1.8))
  ggsave(plot_file, plot = primer_hit_plot, width = 14, height = plot_height, dpi = 150, limitsize = FALSE)
  knitr::include_graphics(plot_file)
} else {
  cat("No compatible primer-end overlaps were detected at the configured minimum overlap of ",
      primer_min_overlap, " nt.", sep = "")
}
```

## Per-sequence input QC

```{r input-per-seq}
if (nrow(input_seq_summary) > 0) {
  input_seq_summary %>%
    transmute(
      Sequence = seqName,
      Length = length,
      Ambiguous = ambiguous_count,
      `Ambiguous bases` = if_else(ambiguous_chars == "", "—", ambiguous_chars),
      `Primer-like count` = primer_like_count,
      `Primer-like bases` = if_else(primer_like_chars == "", "—", primer_like_chars)
    ) %>%
    kbl() %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
} else {
  cat("Input FASTA not found:", input_fasta)
}
```

---

```{r amplicon-coverage, results='asis', fig.width=12}
if (has_amplicon && nrow(amplicon_long) > 0) {
  cat("# Amplicon coverage (whole-genome / ViroConstrictor) {.tabset}\n\n")

  # ── Tab 1: coverage barplot, one bar per amplicon, faceted per sample ───────
  cat("## Coverage barplot\n\n")

  .mincov <- if (!is.na(mincov_val)) mincov_val else 20

  # Per-sample genotype (NextClade clade) for the facet headers + grid ordering.
  .geno_lookup <- triage_tbl %>%
    transmute(sample = word(seqName, 1),
              geno   = harmonise_genotype(clade)) %>%
    distinct(sample, .keep_all = TRUE)

  amp_plot <- amplicon_long %>%
    left_join(.geno_lookup, by = "sample") %>%
    mutate(
      geno     = if_else(is.na(geno), "—", geno),
      passes   = coverage >= .mincov,
      cov_plot = pmax(coverage, 0.5),       # floor so zeros are visible on log scale
      facet_label = paste0(sample, " · ", geno)
    )

  # Order facets by genotype (no-call last), then sample, so same-genotype
  # samples sit together in the grid.
  .facet_levels <- amp_plot %>%
    distinct(sample, geno, facet_label) %>%
    arrange(geno == "—", geno, sample) %>%
    pull(facet_label)
  amp_plot <- amp_plot %>%
    mutate(facet_label = factor(facet_label, levels = .facet_levels))

  n_samples <- dplyr::n_distinct(amp_plot$sample)
  p_amp <- ggplot(amp_plot, aes(x = factor(amplicon), y = cov_plot, fill = passes)) +
    geom_col() +
    geom_hline(yintercept = .mincov, linetype = "dashed", colour = "grey40", linewidth = 0.3) +
    facet_wrap(~ facet_label, ncol = 4) +
    scale_y_log10(labels = scales::label_number()) +
    scale_fill_manual(
      values = c(`TRUE` = "#2ca02c", `FALSE` = "#d62728"),
      labels = c(`TRUE` = paste0("≥ MinCov (", .mincov, "×)"),
                 `FALSE` = paste0("< MinCov (", .mincov, "×)")),
      name = NULL, drop = FALSE
    ) +
    labs(
      x = "Amplicon", y = "Mean coverage (×, log scale)",
      title = "Per-amplicon mean coverage by sample",
      subtitle = paste0("Dashed line = MinCov threshold (", .mincov,
                        "×). Bars below it (red) failed the consensus depth filter.")
    ) +
    theme_bw(base_size = 11) +
    theme(legend.position = "bottom",
          panel.grid.major.x = element_blank())

  .tmp_amp <- tempfile(fileext = ".png")
  ggsave(.tmp_amp, plot = p_amp, width = 12,
         height = 2.4 * ceiling(n_samples / 4) + 1, dpi = 150, limitsize = FALSE)
  cat(sprintf('<img src="%s" style="width:100%%;"/>\n\n', knitr::image_uri(.tmp_amp)))
  file.remove(.tmp_amp)

  # Mean coverage per amplicon across the batch (compact summary table)
  amp_mean <- amplicon_long %>%
    group_by(Amplicon = amplicon) %>%
    summarise(
      `Mean coverage (×)`   = round(mean(coverage, na.rm = TRUE), 1),
      `Median coverage (×)` = round(median(coverage, na.rm = TRUE), 1),
      `Samples ≥ MinCov`    = sum(coverage >= .mincov, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(`Samples ≥ MinCov` = paste0(`Samples ≥ MinCov`, " / ", n_samples))
  cat(kbl(amp_mean, caption = "Per-amplicon coverage across the batch") %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                      full_width = FALSE) %>%
        as.character())
  cat("\n\n")

  # ── Tab 2: primer scheme table (primer name ↔ amplicon number) ──────────────
  cat("## Primer scheme\n\n")
  if (nrow(primer_scheme_tbl) > 0) {
    .ptbl <- primer_scheme_tbl %>%
      select(Scheme = scheme, Amplicon = amplicon,
             `Left primer`            = any_of("primer_LEFT"),
             `Left primer (original)` = any_of("original_LEFT"),
             `Left sequence`          = any_of("sequence_LEFT"),
             `Right primer`            = any_of("primer_RIGHT"),
             `Right primer (original)` = any_of("original_RIGHT"),
             `Right sequence`          = any_of("sequence_RIGHT"))
    cat(kbl(.ptbl, caption = "Primer scheme: primer names (ViroConstrictor & original) and sequences per amplicon") %>%
          kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                        full_width = FALSE) %>%
          as.character())
    cat("\n\n")
  } else {
    cat("*Primer scheme FASTA referenced by the samplesheet could not be read.*\n\n")
  }

  cat("---\n\n")
}
```

# 1. Per-sequence summary {.tabset}

```{r summary-table}
summary_tbl %>%
  mutate(
    coverage       = if_else(!is.na(coverage) & is.numeric(coverage), percent(coverage, 0.1), "—"),
    blast_pident   = if_else(!is.na(blast_pident),   paste0(round(blast_pident, 1), "%"), "—"),
    blast_mismatch = if_else(!is.na(blast_mismatch), as.character(blast_mismatch),        "—"),
    blast_qcovs    = if_else(!is.na(blast_qcovs),    paste0(blast_qcovs, "%"),           "—"),
    clade          = if_else(!is.na(clade),           clade,                               "—"),
    lineage_phylo  = if_else(!is.na(lineage_phylo),   lineage_phylo,                       "—"),
    blast_lineage  = if_else(!is.na(blast_lineage),   blast_lineage,                       "—"),
    blast_genotype = if_else(!is.na(blast_genotype),  blast_genotype,                      "—")
  ) %>%
  select(
    Sequence          = seqName,
    `Clade`           = clade,
    `Lineage (phylo)` = lineage_phylo,
    `NC QC`           = qc_status,
    `BLAST top hit`   = sseqid,
    `BLAST %id`       = blast_pident,
    `BLAST SNPs`      = blast_mismatch,
    `BLAST qcov%`     = blast_qcovs,
    `BLAST lineage`   = blast_lineage
  ) -> summary_display

# Calculate background colors for BLAST SNPs column (7th column)
snp_colors <- summary_display$`BLAST SNPs` %>%
  map_chr(~ {
    if (. == "—") {
      ""  # Empty string for no background
    } else {
      snp_count <- as.numeric(.)
      if (snp_count == 0) {
        "#c8e6c9"  # Light green for 0 SNPs
      } else if (snp_count <= 2) {
        "#fff9c4"  # Light yellow for 1-2 SNPs
      } else {
        ""  # Empty string for no background
      }
    }
  })

summary_display %>%
  kbl(escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
  column_spec(2, bold = TRUE) %>%
  column_spec(3, bold = TRUE, color = "#1a6bb5") %>%
  column_spec(7, bold = TRUE, background = snp_colors)
```

---

# 2. Clade assignment comparison {.tabset}

## Agreement matrix

```{r clade-comparison}
clade_comp <- summary_tbl %>%
  select(seqName, clade, lineage_phylo, blast_genotype, blast_lineage) %>%
  mutate(
    blast_vs_clade = case_when(
      is.na(blast_genotype) | is.na(clade) ~ "No BLAST hit",
      str_detect(
        str_to_upper(clade),
        str_remove(str_to_upper(replace_na(blast_genotype, "")), "[Aa]$|[Bb]$")
      ) ~ "Agreement",
      TRUE ~ "DISCORDANT"
    )
  )

clade_comp %>%
  select(
    Sequence             = seqName,
    `Clade`              = clade,
    `lineage_phylo`      = lineage_phylo,
    `BLAST genotype`     = blast_genotype,
    `BLAST lineage`      = blast_lineage,
    `Clade vs BLAST`     = blast_vs_clade
  ) %>%
  mutate(
    `lineage_phylo`  = if_else(is.na(`lineage_phylo`),  "—", `lineage_phylo`),
    `BLAST genotype` = if_else(is.na(`BLAST genotype`), "—", `BLAST genotype`),
    `BLAST lineage`  = if_else(is.na(`BLAST lineage`),  "—", `BLAST lineage`)
  ) %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
  column_spec(6, color = "white",
              background = if_else(clade_comp$blast_vs_clade == "DISCORDANT", "#d62728",
                           if_else(clade_comp$blast_vs_clade == "Agreement",  "#2ca02c", "#999999")))
```

## Clade vs BLAST genotype

```{r blast-clade-plot, fig.height=4}
clade_comp %>%
  count(clade, blast_genotype, blast_vs_clade) %>%
  mutate(label = paste0(replace_na(clade, "?"), " / ", replace_na(blast_genotype, "?"))) %>%
  ggplot(aes(x = label, y = n, fill = blast_vs_clade)) +
  geom_col() +
  scale_fill_manual(values = c("Agreement" = "#2ca02c", "DISCORDANT" = "#d62728",
                                "No BLAST hit" = "#999999")) +
  labs(x = "Clade / BLAST genotype", y = "Count", fill = "Status",
       title = "Clade vs BLAST genotype agreement") +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

---

# 3. Mutational profiles {.tabset}

## Nucleotide substitutions

```{r mut-profiles}
# Mutation table: one row per sequence showing substitution count and list
summary_tbl %>%
  select(seqName,
         `NC-lineages count` = totalSubstitutions,
         `NC-lineages subs`  = substitutions) %>%
  mutate(across(ends_with("subs"), ~if_else(is.na(.), "—", as.character(.)))) %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(width = "100%", height = "400px")
```

## Amino acid substitutions

```{r aa-subs}
summary_tbl %>%
  select(seqName, clade, lineage_phylo,
         `NC-lineages AA` = aaSubstitutions) %>%
  mutate(across(contains("AA"), ~if_else(is.na(.), "—", .))) %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  scroll_box(width = "100%", height = "400px")
```

## Substitution counts

```{r subs-count-plot, fig.height=5}
if (any(!is.na(summary_tbl$totalSubstitutions))) {
  ggplot(summary_tbl %>% filter(!is.na(totalSubstitutions)),
         aes(x = seqName, y = totalSubstitutions)) +
    geom_point(size = 3, color = "#1a6bb5") +
    geom_segment(aes(xend = seqName, yend = 0), linewidth = 0.5, color = "#1a6bb5") +
    labs(x = "Sequence", y = "Total substitutions vs reference",
         title = "Substitution counts (NC-lineages)") +
    theme_bw(base_size = 13) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}
```

---

# 4. BLAST analysis detail {.tabset}

## Top hits per sequence

```{r blast-detail}
if (nrow(blast_raw) > 0) {
  # Prepare metadata join with safe column selection
  meta_to_join <- meta %>%
    select(id, any_of("date")) %>%
    rename(sseqid = id)
  
  if ("date" %in% names(meta_to_join)) {
    meta_to_join <- meta_to_join %>% rename(hit_date = date)
  }
  
  blast_top %>%
    filter(hit_rank <= 10) %>%
    left_join(meta_to_join, by = "sseqid") %>%
    select(
      Query    = qseqid,
      Rank     = hit_rank,
      Hit      = sseqid,
      `% identity` = pident,
      `Query cov%`  = qcovs,
      `SNPs`        = mismatch,
      `Bitscore`    = bitscore,
      any_of(c("hit_genotype", "hit_lineage", "hit_date"))
    ) %>%
    rename_with(~case_when(
      . == "hit_genotype" ~ "Genotype",
      . == "hit_lineage" ~ "Lineage",
      . == "hit_date" ~ "Date",
      TRUE ~ .
    )) %>%
    mutate(
      `% identity` = round(`% identity`, 2),
      across(any_of(c("Genotype", "Lineage", "Date")), 
             ~replace_na(as.character(.), "—"))
    ) %>%
    kbl() %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
    scroll_box(height = "500px")
} else {
  cat("No BLAST results found.")
}
```

## SNP distance to top hits

```{r blast-snp-plot, fig.height=5}
if (nrow(blast_raw) > 0) {
  blast_top %>%
    filter(hit_rank <= 15) %>%
    mutate(label = paste0(sseqid, "\n", replace_na(hit_lineage, "?"))) %>%
    ggplot(aes(x = reorder(label, mismatch), y = mismatch, color = hit_genotype)) +
    geom_point(size = 2.5) +
    geom_segment(aes(xend = reorder(label, mismatch), yend = 0), linewidth = 0.5) +
    facet_wrap(~qseqid, scales = "free_x") +
    coord_flip() +
    labs(x = NULL, y = "SNP distance (mismatches)", color = "Genotype",
         title = "BLAST: SNP distance to top 15 hits per query") +
    theme_bw(base_size = 11)
}
```

---

# 5. NextClade detail {.tabset}

## Nearest nodes

```{r nc-detail-nodes}
# Parse nearestNodeName from NDJSON
parse_nearest <- function(ndjson_path, run_name) {
  if (!file.exists(ndjson_path)) return(tibble())
  lines <- readLines(ndjson_path)
  map_dfr(lines[nchar(lines) > 2], function(l) {
    tryCatch({
      rec <- fromJSON(l)
      tibble(
        seqName     = rec$seqName     %||% NA_character_,
        run         = run_name,
        nearestNode = rec$nearestNodeName %||% NA_character_
      )
    }, error = function(e) tibble())
  })
}

nodes_lineages <- parse_nearest(file.path(out_dir, "lineages", "nextclade.ndjson"), "NC-lineages")

nodes_lineages %>%
  select(Sequence = seqName, `Nearest node (NC-lineages)` = nearestNode) %>%
  mutate(across(everything(), ~if_else(is.na(.), "—", as.character(.)))) %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

## Private mutations

```{r priv-muts-detail}
summary_tbl %>%
  select(
    Sequence          = seqName,
    `Clade`           = clade,
    `lineage_phylo`   = lineage_phylo,
    `Priv.SNPs`       = privateNucMuts_total,
    `QC`              = qc_status
  ) %>%
  mutate(
    `Clade`         = if_else(is.na(`Clade`),         "—", `Clade`),
    `lineage_phylo` = if_else(is.na(`lineage_phylo`), "—", `lineage_phylo`)
  ) %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

---

# 6. Per-sequence details {.tabset}

<div id="tree-lightbox" onclick="this.style.display='none'" style="display:none;position:fixed;top:0;left:0;width:100%;height:100%;background:rgba(0,0,0,.92);z-index:9999;cursor:zoom-out;overflow:auto;display:none;align-items:center;justify-content:center;">
  <img id="tree-lightbox-img" style="max-width:95vw;max-height:95vh;display:block;margin:auto;"/>
  <div style="position:fixed;top:12px;right:20px;color:#fff;font-size:28px;cursor:pointer;" onclick="document.getElementById('tree-lightbox').style.display='none'">&times;</div>
</div>

## {.unlisted .unnumbered}

```{r per-seq-details, results='asis'}
suppressPackageStartupMessages({
  library(pheatmap)
  library(ggmsa)
})

# ── Shared setup ──────────────────────────────────────────────────────────────
ref_meta <- meta %>%
  select(id, genotype, lineage, lineage_phylo, date) %>%
  mutate(
    id_norm       = norm_id(id),
    genotype      = str_to_upper(genotype),
    lineage_phylo = if_else(is.na(lineage_phylo) | lineage_phylo == "", NA_character_, lineage_phylo)
  )

query_lp <- nc_lineages %>%
  select(seqName, lineage_phylo) %>%
  filter(!is.na(lineage_phylo), is_lineage(lineage_phylo))

query_ids <- summary_tbl$seqName
called    <- c("a", "c", "g", "t")

# Normalize query IDs to match tree directory names (first token only, trimmed)
query_ids_normalized <- word(query_ids, 1)

.is_called <- function(x) {
  x %in% called
}

.consensus_base <- function(col) {
  vals <- col[.is_called(col)]
  if (length(vals) == 0) return(NA_character_)
  tab <- sort(table(vals), decreasing = TRUE)
  names(tab)[1]
}

.col_entropy <- function(col) {
  vals <- col[.is_called(col)]
  if (length(vals) == 0) return(NA_real_)
  p <- as.numeric(table(vals)) / length(vals)
  -sum(p * log2(p))
}

.roll_mean <- function(x, k = 15) {
  n <- length(x)
  if (n == 0) return(numeric(0))
  k <- max(1L, as.integer(k))
  h <- floor(k / 2)
  sapply(seq_len(n), function(i) {
    lo <- max(1L, i - h)
    hi <- min(n, i + h)
    mean(x[lo:hi], na.rm = TRUE)
  })
}

.rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  if (!all(is.finite(rng)) || diff(rng) == 0) return(rep(0, length(x)))
  (x - rng[1]) / diff(rng)
}

.pick_zoom_windows <- function(score, n_windows = 5, width = 35) {
  if (length(score) == 0 || all(!is.finite(score))) {
    return(tibble(start = integer(), end = integer(), center = integer(), score = double()))
  }
  width <- max(5L, as.integer(width))
  half  <- floor(width / 2)
  ord   <- order(score, decreasing = TRUE, na.last = NA)
  chosen <- integer()
  for (idx in ord) {
    if (length(chosen) >= n_windows) break
    if (length(chosen) == 0 || all(abs(idx - chosen) > width)) {
      chosen <- c(chosen, idx)
    }
  }
  if (length(chosen) == 0) {
    return(tibble(start = integer(), end = integer(), center = integer(), score = double()))
  }
  tibble(
    center = chosen,
    start  = pmax(1L, chosen - half),
    end    = pmin(length(score), chosen + half),
    score  = score[chosen]
  ) %>%
    arrange(desc(score))
}

.embed_img <- function(path) {
  cat(sprintf(
    '<img src="%s" onclick="document.getElementById(\'tree-lightbox-img\').src=this.src; var lb=document.getElementById(\'tree-lightbox\'); lb.style.display=\'flex\';" style="width:100%%; cursor:zoom-in; border:1px solid #ddd; border-radius:4px; margin-bottom:12px;" title="Click to enlarge"/>\n',
    knitr::image_uri(path)
  ))
}

treefile_exists <- sapply(query_ids_normalized, function(id) file.exists(file.path(trees_dir, id, "tree.treefile")))
aln_exists <- sapply(query_ids_normalized, function(id) {
  # Prefer BLAST-only alignment if available, otherwise fall back to full alignment
  blast_only <- file.exists(file.path(trees_dir, id, "aligned_trimmed_blast_only.fa"))
  full_aln <- file.exists(file.path(trees_dir, id, "aligned_trimmed.fa"))
  blast_only || full_aln
})

if (length(query_ids) == 0) {
  cat("## Status\n\n")
  cat("*No sequences available in summary table for per-sequence details.*\n\n")
} else if (!any(treefile_exists) && !any(aln_exists)) {
  cat("## Status\n\n")
  cat("Per-sequence tree outputs are not present for this batch, so detailed tree/alignment panels are skipped.\n\n")

  status_tbl <- tibble(
    Sequence = query_ids,
    `Tree file` = if_else(treefile_exists, "Found", "Missing"),
    `Tree alignment` = if_else(aln_exists, "Found", "Missing")
  )

  cat(kbl(status_tbl, caption = "Per-sequence tree artifacts") %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                      full_width = TRUE) %>%
        as.character())
  cat("\n\n")

  cat("Run this first (from project root) to populate section 6:\n\n")
  cat("```bash\n")
  cat(sprintf("bash scripts/build_per_seq_trees.sh %s %s\n", params$batch_dir, params$dataset_date))
  cat("```\n\n")
  cat("Short-region alignment inspection is still available in section 0 under **Junction alignment (focused viewer)**.\n\n")
} else {
for (i in seq_along(query_ids)) {
  query_id <- query_ids[i]
  query_id_norm <- query_ids_normalized[i]
  cat(paste0("\n## ", query_id, "\n\n"))

  tree_dir <- file.path(trees_dir, query_id_norm)
  # Prefer BLAST-only alignment if available, otherwise fall back to full alignment
  aln_file <- if (file.exists(file.path(tree_dir, "aligned_trimmed_blast_only.fa"))) {
    file.path(tree_dir, "aligned_trimmed_blast_only.fa")
  } else {
    file.path(tree_dir, "aligned_trimmed.fa")
  }
  treefile <- file.path(tree_dir, "tree.treefile")
  tr <- NULL

  # Read alignment once — used by both SNP and MSA sections
  seqs_raw <- if (file.exists(aln_file)) {
    tryCatch(ape::read.dna(aln_file, format = "fasta", as.character = TRUE),
             error = function(e) NULL)
  } else NULL

  # ── 1. Phylogenetic tree ────────────────────────────────────────────────────
  cat("### Phylogenetic tree\n\n")
  if (!file.exists(treefile)) {
    cat("*Tree not available — run `build_per_seq_trees.sh` first.*\n\n")
  } else {
    tr <- tryCatch(read.tree(treefile), error = function(e) NULL)
    if (is.null(tr)) {
      cat("*Could not parse tree.*\n\n")
    } else {
      tr <- phangorn::midpoint(tr)

      tip_info <- tibble(id = tr$tip.label) %>%
        left_join(ref_meta %>% select(id_norm, genotype, lineage, lineage_phylo, date),
                  by = c("id" = "id_norm")) %>%
        left_join(query_lp %>% dplyr::rename(id = seqName, lineage_phylo_q = lineage_phylo),
                  by = "id") %>%
        left_join(community_meta, by = "id") %>%
        mutate(
          lineage_phylo = coalesce(lineage_phylo_q, lineage_phylo, community_lineage_phylo),
          genotype      = coalesce(genotype, community_clade),
          is_query  = id == query_id,
          tip_label = case_when(
            !is.na(lineage_phylo) & !is.na(lineage) & lineage != "NA" ~
              paste0(id, "  [", lineage, "  /  ", lineage_phylo, "]"),
            !is.na(lineage_phylo) & !is.na(community_clade) ~
              paste0(id, "  [", community_clade, "  /  ", lineage_phylo, "]"),
            !is.na(lineage_phylo) ~
              paste0(id, "  [", lineage_phylo, "]"),
            !is.na(lineage) & lineage != "NA" ~
              paste0(id, "  [", lineage, "]"),
            !is.na(community_clade) ~
              paste0(id, "  [", community_clade, "]"),
            TRUE ~ id
          ),
          tip_color = case_when(
            is_query                               ~ "Query",
            !is.na(lineage) & lineage != "NA"      ~ lineage,
            !is.na(genotype) & genotype != "NA"    ~ genotype,
            !is.na(community_clade)                ~ community_clade,
            TRUE ~ "Unknown"
          ),
          date_show = if_else(!is.na(date) & date != "NA", date, "—")
        )

      all_colors <- unique(tip_info$tip_color)
      ref_colors <- setdiff(all_colors, "Query")
      pal <- setNames(
        c("#d62728", scales::hue_pal()(length(ref_colors))),
        c("Query", ref_colors)
      )

      p <- ggtree(tr, layout = "rectangular", branch.length = "branch.length") %<+% tip_info +
        geom_tippoint(aes(color = tip_color), size = 3) +
        geom_tiplab(aes(label = tip_label), size = 2.8, hjust = -0.05) +
        scale_color_manual(values = pal, name = "Lineage") +
        guides(color = guide_legend(ncol = 2)) +
        theme_tree2() +
        theme(legend.position = "bottom", plot.margin = margin(5, 5, 5, 5)) +
        hexpand(.6)

      bs_df <- p$data %>%
        filter(!isTip, !is.na(label)) %>%
        mutate(bs_val = suppressWarnings(as.numeric(label))) %>%
        filter(!is.na(bs_val), bs_val >= 50)
      if (nrow(bs_df) > 0) {
        p <- p + geom_text(data = bs_df, aes(x = x, y = y, label = round(bs_val)),
                           inherit.aes = FALSE,
                           size = 2.5, hjust = 1.15, vjust = -0.4, color = "grey35")
      }

      .tmp_png <- tempfile(fileext = ".png")
      ggsave(.tmp_png, plot = p, width = 18, height = 9, dpi = 150)
      .embed_img(.tmp_png)
      file.remove(.tmp_png)
      cat("\n")

      meta_tbl <- tip_info %>%
        arrange(desc(is_query), tip_color, id) %>%
        select(ID = id, Lineage = lineage, Lineage_phylo = lineage_phylo,
               Genotype = genotype, Date = date_show) %>%
        mutate(across(c(Lineage, Lineage_phylo, Genotype),
                      ~if_else(is.na(.) | . %in% c("NA", ""), "—",
                               str_to_upper(as.character(.)))))
      cat(kbl(meta_tbl, caption = paste("Sequences in tree:", query_id)) %>%
            kable_styling(bootstrap_options = c("striped", "condensed"),
                          full_width = FALSE, font_size = 11) %>%
            as.character())
      cat("\n\n")
    }
  }

  # ── 2. SNP distance heatmap ─────────────────────────────────────────────────
  cat("### SNP distance heatmap\n\n")
  
  # Filter sequences: keep only query + BLAST database sequences (exclude NextClade references)
  # If too few filtered sequences, fall back to all sequences
  seqs_heatmap <- if (!is.null(seqs_raw) && nrow(seqs_raw) > 0) {
    # Get all sequence IDs from alignment
    all_ids <- rownames(seqs_raw)
    # Keep only: query (if present) + sequences that are in BLAST database
    blast_ids <- intersect(community_meta$id, all_ids)
    query_ids_keep <- if (query_id %in% all_ids) query_id else character()
    keep_ids <- unique(c(query_ids_keep, blast_ids))
    
    # Use filtered set if we have at least 2 sequences; otherwise fall back to all
    if (length(keep_ids) >= 2) {
      seqs_raw[keep_ids, , drop = FALSE]
    } else if (nrow(seqs_raw) >= 2) {
      seqs_raw  # Fall back to all sequences
    } else {
      NULL
    }
  } else {
    NULL
  }
  
  if (is.null(seqs_heatmap) || nrow(seqs_heatmap) < 2) {
    cat("*Alignment not available or too few sequences.*\n\n")
  } else {
    n_tips  <- nrow(seqs_heatmap)
    ids     <- rownames(seqs_heatmap)
    snp_mat <- matrix(0L, n_tips, n_tips, dimnames = list(ids, ids))
    for (i in seq_len(n_tips - 1)) {
      for (j in (i + 1):n_tips) {
        s1 <- seqs_heatmap[i, ]; s2 <- seqs_heatmap[j, ]
        valid <- s1 %in% called & s2 %in% called
        d <- sum(s1[valid] != s2[valid])
        snp_mat[i, j] <- snp_mat[j, i] <- d
      }
    }

    lbl_tbl <- tibble(id = ids) %>%
      left_join(community_meta, by = "id") %>%
      mutate(
        lineage_phylo = community_lineage_phylo,
        lineage       = community_clade,
        lineage_phylo = if_else(is.na(lineage_phylo) | lineage_phylo == "",
                                NA_character_, lineage_phylo),
        # Shorten long FASTA headers: extract accession + isolate/strain name
        # Handles both "isolate XXXX" and "strain: XXXX" formats
        # E.g., "KX151409.1 Hepatovirus A isolate h737 ..." → "KX151409.1 (h737)"
        # E.g., "LC035017.1 Hepatitis A virus gene for polyprotein, partial cds, strain: 1404-20055-WakymC-4HA42" → "LC035017.1 (1404-20055-WakymC-4HA42)"
        short_id = {
          accession <- str_extract(id, "^[^\\s]+")
          # Extract isolate/strain name: try "isolate" pattern, then "strain:" pattern
          isolate_m1 <- str_extract(id, "(?<= isolate )[^ ,]+")
          isolate_m2 <- str_extract(id, "(?<=strain: )[^ ,]+")
          isolate_part <- coalesce(isolate_m1, isolate_m2)
          
          if_else(!is.na(isolate_part),
                 paste0(accession, " (", isolate_part, ")"),
                 accession)
        },
        display = case_when(
          id == query_id ~ paste0(short_id, "  \u2605"),
          !is.na(lineage_phylo) & !is.na(lineage) & lineage != "NA" ~
            paste0(short_id, "  [", lineage, "  /  ", lineage_phylo, "]"),
          !is.na(lineage_phylo) ~ paste0(short_id, "  [", lineage_phylo, "]"),
          !is.na(lineage) & lineage != "NA" ~ paste0(short_id, "  [", lineage, "]"),
          TRUE ~ short_id
        )
      )
    lbl_map <- setNames(lbl_tbl$display, lbl_tbl$id)
    rownames(snp_mat) <- lbl_map[rownames(snp_mat)]
    colnames(snp_mat) <- lbl_map[colnames(snp_mat)]

    query_lbl <- lbl_map[query_id]
    ann_row <- data.frame(
      Type = ifelse(rownames(snp_mat) == query_lbl, "Query", "Reference"),
      row.names = rownames(snp_mat)
    )
    ann_colors <- list(Type = c(Query = "#d62728", Reference = "#aec7e8"))

    # Adaptive sizing: base cell size on number of sequences
    # Ensures readable labels and numbers even with many sequences
    cell_size <- if (n_tips <= 10) 60 else if (n_tips <= 20) 45 else 30
    fontsize_main <- if (n_tips <= 10) 15 else if (n_tips <= 20) 14 else 13
    fontsize_num <- if (n_tips <= 10) 12 else if (n_tips <= 20) 11 else 10
    fontsize_row <- if (n_tips <= 15) 13 else 11
    
    # Calculate figure size with minimum/maximum constraints
    px <- min(18, max(10, n_tips * (cell_size / 35)))
    
    # Continuous blue palette for all SNP distances
    color_func <- colorRampPalette(c("#f7fbff", "#deebf7", "#9ecae1", "#3182bd", "#08519c", "#08306b"))(100)
    
    .tmp_heat <- tempfile(fileext = ".png")
    .ph_obj <- pheatmap(
      snp_mat,
      # Color palette: continuous blue
      color            = color_func,
      clustering_distance_rows = as.dist(snp_mat),
      clustering_distance_cols = as.dist(snp_mat),
      clustering_method        = "average",
      display_numbers          = TRUE,
      number_format            = "%d",
      number_color             = "black",      # Improved contrast
      fontsize                 = fontsize_main,
      fontsize_number          = fontsize_num,
      fontsize_row             = fontsize_row,
      annotation_row           = ann_row,
      annotation_colors        = ann_colors,
      legend                   = TRUE,
      main                     = paste("Pairwise SNP distances \u2014", query_id),
      treeheight_row           = 100,
      treeheight_col           = 100,
      border_color             = "#cccccc",    # Subtle borders for better separation
      gaps_row                 = which(rownames(snp_mat) == query_lbl),
      gaps_col                 = which(colnames(snp_mat) == query_lbl),
      silent                   = TRUE
    )
    ggsave(.tmp_heat, plot = .ph_obj$gtable, width = px, height = px, dpi = 150, limitsize = FALSE)
    .embed_img(.tmp_heat)
    file.remove(.tmp_heat)
    
    # Save SNP distance matrix as TSV for external analysis
    snp_tsv <- file.path(out_dir, "snp_matrices", paste0(query_id, "_snp_distances.tsv"))
    dir.create(dirname(snp_tsv), showWarnings = FALSE, recursive = TRUE)
    
    # Convert matrix to data frame with sequence IDs in first column
    snp_df <- as.data.frame(snp_mat)
    snp_df <- tibble::rownames_to_column(snp_df, var = "Sequence_1")
    names(snp_df)[-1] <- rownames(snp_mat)  # Ensure column names are sequence labels
    
    write_tsv(snp_df, snp_tsv)
    cat("[Download SNP distance matrix](snp_matrices/", query_id, "_snp_distances.tsv)\n\n", sep = "")
    
    cat("\n\n")
  }

  # ── 3. Raw MSA overview (ggmsa) ───────────────────────────────────────────
  cat("### Raw multiple sequence alignment (overview)\n\n")
  if (is.null(seqs_raw) || nrow(seqs_raw) < 2) {
    cat("*Alignment not available or too few sequences.*\n\n")
  } else if (ncol(seqs_raw) > 2000) {
    cat("*Tree alignment is long (", ncol(seqs_raw), " columns). For readable alignment inspection, use the focused junction viewer in section 0 (\"Junction alignment\").*\n\n", sep = "")
  } else {
    mat0 <- tolower(seqs_raw)
    ids0 <- rownames(mat0)

    snp_raw0 <- matrix(0L, nrow(mat0), nrow(mat0), dimnames = list(ids0, ids0))
    if (nrow(mat0) > 1) {
      for (i in seq_len(nrow(mat0) - 1)) {
        for (j in (i + 1):nrow(mat0)) {
          s1 <- mat0[i, ]; s2 <- mat0[j, ]
          valid <- .is_called(s1) & .is_called(s2)
          d <- sum(s1[valid] != s2[valid])
          snp_raw0[i, j] <- snp_raw0[j, i] <- d
        }
      }
    }

    if (!is.null(tr) && all(ids0 %in% tr$tip.label)) {
      seq_order0 <- tr$tip.label[tr$tip.label %in% ids0]
    } else if (length(ids0) > 1) {
      seq_order0 <- tryCatch({
        ids0[hclust(as.dist(snp_raw0[ids0, ids0, drop = FALSE]), method = "average")$order]
      }, error = function(e) ids0)
    } else {
      seq_order0 <- ids0
    }
    seq_order0 <- unique(c(seq_order0, setdiff(ids0, seq_order0)))

    lbl_tbl0 <- tibble(id = seq_order0) %>%
      left_join(ref_meta %>% select(id_norm, lineage, lineage_phylo), by = c("id" = "id_norm")) %>%
      left_join(community_meta, by = "id") %>%
      mutate(
        lineage_phylo = coalesce(lineage_phylo, community_lineage_phylo),
        lineage       = coalesce(lineage, community_clade),
        # Shorten long FASTA headers: extract accession + isolate/strain name
        short_id = {
          accession <- str_extract(id, "^[^\\s]+")
          # Extract isolate/strain name: try "isolate" pattern, then "strain:" pattern
          isolate_m1 <- str_extract(id, "(?<= isolate )[^ ,]+")
          isolate_m2 <- str_extract(id, "(?<=strain: )[^ ,]+")
          isolate_part <- coalesce(isolate_m1, isolate_m2)
          
          if_else(!is.na(isolate_part),
                 paste0(accession, " (", isolate_part, ")"),
                 accession)
        },
        display = case_when(
          id == query_id & !is.na(lineage_phylo) ~ paste0(short_id, " ★ [", lineage_phylo, "]"),
          id == query_id ~ paste0(short_id, " ★"),
          !is.na(lineage_phylo) ~ paste0(short_id, " [", lineage_phylo, "]"),
          !is.na(lineage) ~ paste0(short_id, " [", lineage, "]"),
          TRUE ~ short_id
        )
      )
    lbl_map0 <- setNames(lbl_tbl0$display, lbl_tbl0$id)

    dna_ss0 <- tryCatch(Biostrings::readDNAStringSet(aln_file), error = function(e) NULL)
    if (is.null(dna_ss0)) {
      cat("*Could not read alignment as DNAStringSet.*\n\n")
    } else {
      keep_ids0 <- seq_order0[seq_order0 %in% names(dna_ss0)]
      if (length(keep_ids0) >= 2) {
        dna_ss0 <- dna_ss0[keep_ids0]
        names(dna_ss0) <- lbl_map0[names(dna_ss0)]

        p_raw_msa <- ggmsa(
          dna_ss0,
          start = 1,
          end = Biostrings::width(dna_ss0)[1],
          color = "Chemistry_NT",
          seq_name = TRUE,
          font = NULL,
          show.legend = TRUE
        ) +
          labs(title = paste("Raw MSA overview —", query_id), x = "Alignment position", y = NULL) +
          theme_bw(base_size = 9) +
          theme(
            plot.title = element_text(face = "bold", size = 10),
            axis.text.y = element_text(size = 6.5),
            axis.text.x = element_text(size = 7)
          )

        .tmp_raw_msa <- tempfile(fileext = ".png")
        fig_h_raw <- max(3.2, 0.22 * length(dna_ss0) + 1.8)
        ggsave(.tmp_raw_msa, plot = p_raw_msa, width = 16, height = fig_h_raw, dpi = 150, limitsize = FALSE)
        .embed_img(.tmp_raw_msa)
        file.remove(.tmp_raw_msa)
        cat("*Figure legend:* Full alignment overview with original nucleotide coloring. Rows are ordered by phylogenetic tips when available, otherwise SNP-distance clustering.\n\n")
      } else {
        cat("*Not enough ordered sequences available for raw MSA overview.*\n\n")
      }
    }
  }

  # ── 4. Multiple sequence alignment diagnostics ─────────────────────────────
  cat("### Alignment diagnostics (multi-panel)\n\n")
  if (is.null(seqs_raw) || nrow(seqs_raw) < 2) {
    cat("*Alignment not available or too few sequences.*\n\n")
  } else if (ncol(seqs_raw) > 2000) {
    cat("*Diagnostics for full tree alignments are intentionally suppressed for long alignments (", ncol(seqs_raw), " columns). Use section 0 focused junction alignment plots for inspection of base-level patterns.*\n\n", sep = "")
  } else {
    mat <- tolower(seqs_raw)
    if (is.null(colnames(mat))) {
      colnames(mat) <- as.character(seq_len(ncol(mat)))
    }

    ids <- rownames(mat)
    n_seqs <- nrow(mat)
    pos_all <- seq_len(ncol(mat))

    min_called_fraction <- params$min_called_fraction %||% 0.5
    zoom_count <- as.integer(params$zoom_count %||% 5)
    zoom_width <- as.integer(params$zoom_width %||% 35)
    entropy_window <- as.integer(params$entropy_window %||% 15)

    # Prefer phylogenetic tip order when available; fallback to SNP clustering.
    snp_raw <- matrix(0L, n_seqs, n_seqs, dimnames = list(ids, ids))
    for (i in seq_len(n_seqs - 1)) {
      for (j in (i + 1):n_seqs) {
        s1 <- mat[i, ]; s2 <- mat[j, ]
        valid <- .is_called(s1) & .is_called(s2)
        d <- sum(s1[valid] != s2[valid])
        snp_raw[i, j] <- snp_raw[j, i] <- d
      }
    }

    if (!is.null(tr) && all(ids %in% tr$tip.label)) {
      seq_order <- tr$tip.label[tr$tip.label %in% ids]
    } else if (length(ids) > 1) {
      seq_order <- tryCatch({
        ids[hclust(as.dist(snp_raw[ids, ids, drop = FALSE]), method = "average")$order]
      }, error = function(e) ids)
    } else {
      seq_order <- ids
    }
    seq_order <- unique(c(seq_order, setdiff(ids, seq_order)))
    mat_ord <- mat[seq_order, , drop = FALSE]

    label_tbl <- tibble(id = seq_order) %>%
      left_join(ref_meta %>% select(id_norm, lineage, lineage_phylo), by = c("id" = "id_norm")) %>%
      left_join(community_meta, by = "id") %>%
      mutate(
        lineage_phylo = coalesce(lineage_phylo, community_lineage_phylo),
        lineage       = coalesce(lineage, community_clade),
        # Shorten long FASTA headers: extract accession + isolate/strain name
        short_id = {
          accession <- str_extract(id, "^[^\\s]+")
          # Extract isolate/strain name: try "isolate" pattern, then "strain:" pattern
          isolate_m1 <- str_extract(id, "(?<= isolate )[^ ,]+")
          isolate_m2 <- str_extract(id, "(?<=strain: )[^ ,]+")
          isolate_part <- coalesce(isolate_m1, isolate_m2)
          
          if_else(!is.na(isolate_part),
                 paste0(accession, " (", isolate_part, ")"),
                 accession)
        },
        display = case_when(
          id == query_id & !is.na(lineage_phylo) ~ paste0(short_id, " ★ [", lineage_phylo, "]"),
          id == query_id ~ paste0(short_id, " ★"),
          !is.na(lineage_phylo) ~ paste0(short_id, " [", lineage_phylo, "]"),
          !is.na(lineage) ~ paste0(short_id, " [", lineage, "]"),
          TRUE ~ short_id
        )
      )
    lbl_map <- setNames(label_tbl$display, label_tbl$id)
    disp_levels <- rev(lbl_map[seq_order])

    col_called_n <- apply(mat_ord, 2, function(col) sum(.is_called(col)))
    col_called_frac <- col_called_n / nrow(mat_ord)
    keep_cov <- col_called_frac >= min_called_fraction

    if (!any(keep_cov)) {
      cat("*No columns passed minimum called-site coverage.*\n\n")
    } else {
      mat_cov <- mat_ord[, keep_cov, drop = FALSE]
      pos_cov <- pos_all[keep_cov]

      consensus <- apply(mat_cov, 2, .consensus_base)
      entropy <- apply(mat_cov, 2, .col_entropy)
      gap_density <- colMeans(mat_cov == "-")
      ambig_density <- apply(mat_cov, 2, function(col) mean(!(col %in% c(called, "-"))))
      conservation <- sapply(seq_len(ncol(mat_cov)), function(k) {
        col <- mat_cov[, k]
        vals <- col[.is_called(col)]
        if (length(vals) == 0) return(NA_real_)
        max(as.numeric(table(vals))) / length(vals)
      })
      snp_count <- sapply(seq_len(ncol(mat_cov)), function(k) {
        col <- mat_cov[, k]
        cons <- consensus[k]
        if (is.na(cons)) return(0)
        sum(.is_called(col) & col != cons)
      })
      snp_density <- .roll_mean(snp_count, k = entropy_window)
      var_cols <- which(snp_count > 0)
      var_pos <- pos_cov[var_cols]

      # ── Panel A: Global tracks ─────────────────────────────────────────────
      cat("**Panel A. Global alignment tracks**\n\n")
      cat("*Figure legend:* Entropy, rolling SNP density, gap density, and conservation across alignment coordinates.\n\n")
      track_df <- tibble(
        position = pos_cov,
        Entropy = entropy,
        `SNP density (rolling)` = snp_density,
        `Gap density` = gap_density,
        Conservation = conservation
      ) %>%
        pivot_longer(-position, names_to = "metric", values_to = "value")

      p_track <- ggplot(track_df, aes(x = position, y = value)) +
        geom_line(color = "#2c3e50", linewidth = 0.4) +
        facet_grid(metric ~ ., scales = "free_y", switch = "y") +
        labs(x = "Alignment position", y = NULL,
             title = paste("Alignment signal overview —", query_id)) +
        theme_bw(base_size = 11) +
        theme(
          strip.background = element_rect(fill = "#f5f5f5", color = NA),
          strip.placement = "outside",
          panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold", size = 11)
        )
      .tmp_track <- tempfile(fileext = ".png")
      ggsave(.tmp_track, plot = p_track, width = 16, height = 6.8, dpi = 150, limitsize = FALSE)
      .embed_img(.tmp_track)
      file.remove(.tmp_track)

      # ── Panel B: Variant-only alignment ───────────────────────────────────
      cat("**Panel B. Variant-only alignment**\n\n")
      cat("*Figure legend:* Only variable sites are shown; invariant columns are removed to emphasize informative differences.\n\n")
      if (length(var_cols) == 0) {
        cat("*No variable columns after coverage filtering.*\n\n")
      } else {
        mat_var <- mat_cov[, var_cols, drop = FALSE]
        var_grid <- expand.grid(
          seq_id = rownames(mat_var),
          col_i = seq_len(ncol(mat_var)),
          KEEP.OUT.ATTRS = FALSE,
          stringsAsFactors = FALSE
        ) %>%
          as_tibble() %>%
          mutate(
            base = mat_var[cbind(match(seq_id, rownames(mat_var)), col_i)],
            position = var_pos[col_i],
            seq_label = lbl_map[seq_id],
            base_class = case_when(
              base %in% called ~ str_to_upper(base),
              base == "-" ~ "Gap",
              TRUE ~ "Ambig"
            )
          )

        nt_pal <- c(
          "A" = "#4daf4a", "C" = "#377eb8", "G" = "#ff7f00", "T" = "#e41a1c",
          "Gap" = "#f5f5f5", "Ambig" = "#6a3d9a"
        )

        p_var <- ggplot(var_grid,
                        aes(x = position, y = factor(seq_label, levels = disp_levels), fill = base_class)) +
          geom_tile(color = "white", linewidth = 0.15) +
          scale_fill_manual(values = nt_pal, drop = FALSE) +
          labs(
            x = "Variable alignment position",
            y = NULL,
            fill = "Base",
            title = sprintf("Variant-only view (%d variable sites)", length(var_cols))
          ) +
          theme_bw(base_size = 10) +
          theme(
            axis.text.y = element_text(size = 7),
            panel.grid = element_blank(),
            plot.title = element_text(face = "bold", size = 10),
            legend.position = "bottom"
          )

        .tmp_var <- tempfile(fileext = ".png")
        fig_h_var <- max(3.5, 0.24 * nrow(mat_var) + 1.8)
        ggsave(.tmp_var, plot = p_var, width = 16, height = fig_h_var, dpi = 150, limitsize = FALSE)
        .embed_img(.tmp_var)
        file.remove(.tmp_var)
      }

      # ── Panel C: Consensus-difference matrix ──────────────────────────────
      cat("**Panel C. Consensus-difference matrix**\n\n")
      cat("*Figure legend:* Each tile is coded relative to consensus, or the majority nucleotide, per column (consensus match, difference, gap, or ambiguous call).\n\n")
      cons_grid <- expand.grid(
        seq_id = rownames(mat_cov),
        col_i = seq_len(ncol(mat_cov)),
        KEEP.OUT.ATTRS = FALSE,
        stringsAsFactors = FALSE
      ) %>%
        as_tibble() %>%
        mutate(
          base = mat_cov[cbind(match(seq_id, rownames(mat_cov)), col_i)],
          cons = consensus[col_i],
          position = pos_cov[col_i],
          seq_label = lbl_map[seq_id],
          state = case_when(
            base == "-" ~ "Gap",
            !(base %in% c(called, "-")) ~ "Ambig",
            is.na(cons) ~ "No consensus",
            base == cons ~ "Consensus",
            TRUE ~ "Diff"
          )
        )

      diff_pal <- c(
        "Consensus" = "#e5e7eb",
        "Diff" = "#0f766e",
        "Gap" = "#ffffff",
        "Ambig" = "#4c1d95",
        "No consensus" = "#d1d5db"
      )

      p_cons <- ggplot(cons_grid,
                       aes(x = position, y = factor(seq_label, levels = disp_levels), fill = state)) +
        geom_tile(color = "white", linewidth = 0.1) +
        scale_fill_manual(values = diff_pal, drop = FALSE) +
        labs(
          x = "Alignment position",
          y = NULL,
          fill = "State",
          title = "Consensus-relative map"
        ) +
        theme_bw(base_size = 10) +
        theme(
          axis.text.y = element_text(size = 7),
          panel.grid = element_blank(),
          plot.title = element_text(face = "bold", size = 10),
          legend.position = "bottom"
        )

      .tmp_cons <- tempfile(fileext = ".png")
      fig_h_cons <- max(3.5, 0.24 * nrow(mat_cov) + 1.8)
      ggsave(.tmp_cons, plot = p_cons, width = 16, height = fig_h_cons, dpi = 150, limitsize = FALSE)
      .embed_img(.tmp_cons)
      file.remove(.tmp_cons)

      # ── Panel D: Sequence QC strip ────────────────────────────────────────
      cat("**Panel D. Sequence QC track**\n\n")
      cat("*Figure legend:* Per-sequence quality indicators: private SNPs, ambiguous-base fraction, gap fraction, and distance to consensus.\n\n")
      private_counts <- setNames(rep(0L, nrow(mat_cov)), rownames(mat_cov))
      if (length(var_cols) > 0) {
        for (k in var_cols) {
          col <- mat_cov[, k]
          cons <- consensus[k]
          called_mask <- .is_called(col)
          if (sum(called_mask) < 2 || is.na(cons)) next
          tab <- table(col[called_mask])
          uniq_alt <- setdiff(names(tab)[tab == 1], cons)
          if (length(uniq_alt) == 0) next
          for (b in uniq_alt) {
            idx <- which(col == b)
            private_counts[idx] <- private_counts[idx] + 1L
          }
        }
      }

      dist_to_consensus <- sapply(seq_len(nrow(mat_cov)), function(i) {
        b <- mat_cov[i, ]
        valid <- .is_called(b) & .is_called(consensus)
        sum(b[valid] != consensus[valid])
      })

      qc_tbl <- tibble(
        seq_id = rownames(mat_cov),
        seq_label = lbl_map[rownames(mat_cov)],
        private_snps = as.numeric(private_counts),
        ambiguous_frac = apply(mat_ord[rownames(mat_cov), , drop = FALSE], 1,
                               function(row) mean(!(row %in% c(called, "-")))),
        gap_frac = rowMeans(mat_ord[rownames(mat_cov), , drop = FALSE] == "-"),
        dist_consensus = as.numeric(dist_to_consensus)
      )

      qc_long <- qc_tbl %>%
        select(seq_label, private_snps, ambiguous_frac, gap_frac, dist_consensus) %>%
        pivot_longer(-seq_label, names_to = "metric", values_to = "value") %>%
        group_by(metric) %>%
        mutate(value_scaled = .rescale01(value)) %>%
        ungroup() %>%
        mutate(
          metric = recode(metric,
                          private_snps = "Private SNPs",
                          ambiguous_frac = "Ambiguous fraction",
                          gap_frac = "Gap fraction",
                          dist_consensus = "Distance to consensus"),
          label = case_when(
            metric %in% c("Ambiguous fraction", "Gap fraction") ~ percent(value, 0.1),
            TRUE ~ as.character(round(value, 1))
          )
        )

      p_qc <- ggplot(qc_long,
                     aes(x = metric, y = factor(seq_label, levels = disp_levels), fill = value_scaled)) +
        geom_tile(color = "white", linewidth = 0.2) +
        geom_text(aes(label = label), size = 2.3, color = "#111827") +
        scale_fill_gradient(low = "#f8fafc", high = "#b91c1c") +
        labs(x = NULL, y = NULL, fill = "Scaled", title = "Per-sequence QC metrics") +
        theme_bw(base_size = 10) +
        theme(
          axis.text.x = element_text(angle = 20, hjust = 1),
          axis.text.y = element_text(size = 7),
          panel.grid = element_blank(),
          plot.title = element_text(face = "bold", size = 10)
        )

      .tmp_qc <- tempfile(fileext = ".png")
      fig_h_qc <- max(3.5, 0.24 * nrow(mat_cov) + 1.8)
      ggsave(.tmp_qc, plot = p_qc, width = 14, height = fig_h_qc, dpi = 150, limitsize = FALSE)
      .embed_img(.tmp_qc)
      file.remove(.tmp_qc)

      # Algorithmic QC annotations.
      cat("**Automatic QC flags**\n\n")
      flags <- character(0)
      if (any(qc_tbl$private_snps >= max(2, quantile(qc_tbl$private_snps, 0.9, na.rm = TRUE)))) {
        flagged <- qc_tbl %>%
          filter(private_snps >= max(2, quantile(private_snps, 0.9, na.rm = TRUE))) %>%
          pull(seq_label)
        flags <- c(flags, paste0("High singleton/private SNP burden: ", paste(flagged, collapse = ", ")))
      }
      if (any(qc_tbl$ambiguous_frac > 0.02)) {
        flagged <- qc_tbl %>% filter(ambiguous_frac > 0.02) %>% pull(seq_label)
        flags <- c(flags, paste0("Elevated ambiguous bases (>2%): ", paste(flagged, collapse = ", ")))
      }
      if (any(gap_density > 0.2)) {
        hot <- pos_cov[which(gap_density > 0.2)]
        flags <- c(flags, paste0("Gap hotspot columns (>20% gaps): ", paste(head(hot, 12), collapse = ", ")))
      }
      if (length(flags) == 0) {
        cat("- No strong artifact flags triggered by default thresholds.\n\n")
      } else {
        for (f in flags) cat("- ", f, "\n", sep = "")
        cat("\n")
      }

      # ── Panel E: Top informative zoom windows ─────────────────────────────
      cat("**Panel E. Zoomed informative windows**\n\n")
      cat("*Figure legend:* Top informative windows selected by combined entropy/SNP/gap signal; tiles show consensus-relative state and nucleotide letter.\n\n")
      score <- .rescale01(replace_na(entropy, 0)) +
        .rescale01(snp_density) +
        .rescale01(gap_density)
      zoom_tbl <- .pick_zoom_windows(score, n_windows = zoom_count, width = zoom_width)

      if (nrow(zoom_tbl) == 0) {
        cat("*No informative windows detected.*\n\n")
      } else {
        for (w in seq_len(nrow(zoom_tbl))) {
          i_start <- zoom_tbl$start[w]
          i_end   <- zoom_tbl$end[w]
          idx <- seq.int(i_start, i_end)

          cat(sprintf("Window %d: positions %d–%d\n\n", w, pos_cov[i_start], pos_cov[i_end]))

          zoom_grid <- expand.grid(
            seq_id = rownames(mat_cov),
            col_i = idx,
            KEEP.OUT.ATTRS = FALSE,
            stringsAsFactors = FALSE
          ) %>%
            as_tibble() %>%
            mutate(
              base = mat_cov[cbind(match(seq_id, rownames(mat_cov)), col_i)],
              cons = consensus[col_i],
              position = pos_cov[col_i],
              seq_label = lbl_map[seq_id],
              state = case_when(
                base == "-" ~ "Gap",
                !(base %in% c(called, "-")) ~ "Ambig",
                is.na(cons) ~ "No consensus",
                base == cons ~ "Consensus",
                TRUE ~ "Diff"
              ),
              base_label = case_when(
                base == "-" ~ "-",
                base %in% called ~ str_to_upper(base),
                TRUE ~ "N"
              ),
              text_color = case_when(
                state %in% c("Diff", "Ambig") ~ "#ffffff",
                TRUE ~ "#111827"
              )
            )

          p_zoom <- ggplot(zoom_grid,
                           aes(x = position, y = factor(seq_label, levels = disp_levels), fill = state)) +
            geom_tile(color = "white", linewidth = 0.12) +
            geom_text(aes(label = base_label, color = text_color), size = 2.0) +
            scale_fill_manual(values = diff_pal, drop = FALSE) +
            scale_color_identity() +
            labs(x = "Alignment position", y = NULL, fill = "State") +
            theme_bw(base_size = 9) +
            theme(
              axis.text.y = element_text(size = 6.5),
              panel.grid = element_blank(),
              legend.position = "none"
            )

          .tmp_zoom <- tempfile(fileext = ".png")
          fig_h_zoom <- max(2.8, 0.22 * nrow(mat_cov) + 1.6)
          ggsave(.tmp_zoom, plot = p_zoom, width = 14, height = fig_h_zoom, dpi = 150, limitsize = FALSE)
          .embed_img(.tmp_zoom)
          file.remove(.tmp_zoom)
        }
        cat("\n")
      }
    }
  }
}
}
```

---

# 7. Conclusions and recommendations

```{r conclusions}
n_seqs        <- nrow(summary_tbl)
n_lp          <- sum(is_lineage(summary_tbl$lineage_phylo), na.rm = TRUE)
n_blast_agree <- sum(summary_tbl$clade_agree_blast == TRUE, na.rm = TRUE)
```

- **`r n_seqs`** sequences analyzed.
- **Lineage resolution:** `r n_lp`/`r n_seqs` sequences assigned a `lineage_phylo` by the lineages dataset.
- **BLAST vs clade agreement:** `r n_blast_agree`/`r sum(!is.na(summary_tbl$clade_agree_blast))` sequences where genotype is available.

---

# 8. Output files

```{r output-files}
tibble(
  File = c(
    file.path(params$out_base, "blast_results.tsv"),
    file.path(params$out_base, "lineages", "nextclade.tsv"),
    file.path(params$out_base, "trees", "<seqName>", "tree.treefile"),
    file.path(params$out_base, "trees", "<seqName>", "aligned_trimmed.fa"),
    file.path(params$out_base, "trees", "<seqName>", "aligned_trimmed_blast_only.fa"),
    file.path(params$out_base, "snp_matrices", "<seqName>_snp_distances.tsv"),
    file.path(params$out_base, "summary.tsv"),
    file.path(params$out_base, "comparison.tsv")
  ),
  Description = c(
    "BLAST hits vs local HAV database",
    "NextClade: lineages dataset (clade + lineage_phylo)",
    "IQ-TREE ML tree per query sequence",
    "MAFFT alignment (query + BLAST hits + NextClade community sequences)",
    "MAFFT alignment with BLAST hits only (no NextClade) — used for heatmap in report",
    "Pairwise SNP distance matrix per query sequence (TSV format for external analysis)",
    "Per-sequence summary table (this report, TSV)",
    "Clade comparison table (this report, TSV)"
  )
) %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = TRUE)
```

```{r write-outputs}
# Write summary TSV
write_tsv(summary_tbl, file.path(params$out_base, "summary.tsv"))

# Write comparison TSV
clade_comp_out <- clade_comp %>%
  left_join(summary_tbl %>% select(seqName, blast_lineage, blast_mismatch = blast_mismatch,
                                    blast_pident, blast_qcovs), by = "seqName")
write_tsv(clade_comp_out, file.path(out_dir, "comparison.tsv"))
```

*Summary written to `r file.path(params$out_base, "summary.tsv")`*  
*Comparison written to `r file.path(params$out_base, "comparison.tsv")`*

---

*Report generated: `r Sys.time()`*  
*Batch: `r batch_name`*  
*Dataset: `r params$dataset_date`*  
